---
title: "DHR clinical variables GEE v04"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 7
    fig_height: 5
    theme: cerulean
    toc: yes
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# Analysis of the longitudinal outcome for the clinical variables

The goal of this analysis is to follow longitudinally the values of the clinical variables and visualize the fraction of variables that moved from the out-of-range value to the reference range and vice versa. Only visits 2-5 will be considered as data feedback occurred only at visit 2, therefore no change should be expected at between visit 1 and 2.

Require the libraries

```{r}
require(openxlsx)
require(tidyverse)
require(geepack)
require(doParallel)
require(doBy)
require(multcomp)
require(lme4)
require(car)
require(boot)
```

Load the data

```{r}
load("/home/francesco.marabita/Projects/DHR/data/Clinical/Clin_data_2018-10-09_newVarNames.RData")
ref.val <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Clinical/Reference_values_updated.xlsx", sheet = 1)
var.cat <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Clinical/Variables_names.xlsx", sheet = 1)
```

Select only numerical measurements, ID variables and make long format. Make sure that observations are ordered by Visit.number as GEE requires that longitudinal observation are ordered by time.

```{r}

sel.var <- c("IID", "ID", "Visit.number", "Age", "Visit.date", "Sex",
             "SystolicBP_mean", "DiastolicBP_mean", "PULSE_mean",
             "BMI", "Waist", "Hip",  "MCV", "MCH",
             "fBLeuk", "BHb", "ERDW", "BHKR", "BTrom","BEryt",
             "STSH",  "SD25", "hsCRP", "PGT", "fPGluk", "SInsu",
             "fPTrigly","fPKol","fPKolHDL", "fpKolLDL","KolHDLrat",     
             "fSLipoA1", "fSLipoB", "ApoB_ApoA1ratio")

clin.data %>%
  filter(Visit.number>=2) %>% 
  select_at(sel.var) %>%
  gather(key="variable", value = "value", -(IID:Sex)) %>% 
  arrange(IID, Visit.number, variable) -> clin.data.long

clin.data %>%
  filter(Visit.number>=2) %>% 
  select_at(sel.var[!sel.var %in% c("IID", "ID", "Visit.number", "Age", "Visit.date", "Sex")]) %>%
  t() -> clin.m
colnames(clin.m) <- clin.data$ID[clin.data$Visit.number>=2]

```

Process reference value table, by including only the selected numerical variables and change the values to numeric

```{r}
ref.val <- ref.val[ref.val$feature.ID %in% rownames(clin.m),]
ref.val <- ref.val[order(match(ref.val$feature.ID, rownames(clin.m))),]

ref.val$ref_low_F <- as.numeric(ref.val$ref_low_F)
ref.val$ref_high_F <- as.numeric(ref.val$ref_high_F)
ref.val$ref_low_M <- as.numeric(ref.val$ref_low_M)
ref.val$ref_high_M <- as.numeric(ref.val$ref_high_M)
```

Create the logical matrix `ref.out`, in a way that `ref.out[i,j]` is T if values are out-of-range.

```{r}
ref.out <- matrix(NA, nrow=nrow(clin.m), ncol=ncol(clin.m),
                  dimnames = list(rownames(clin.m), colnames(clin.m)))
for(i in 1:nrow(clin.m)){
  for(j in 1:ncol(clin.m)){
    if(clin.data$Sex[clin.data$ID==colnames(clin.m)[j]]=="F"){
      ref.out[i,j] <- clin.m[i,j] < ref.val$ref_low_F[i] | clin.m[i,j] > ref.val$ref_high_F[i]
    } else if(clin.data$Sex[clin.data$ID==colnames(clin.m)[j]]=="M"){
      ref.out[i,j] <- clin.m[i,j] < ref.val$ref_low_M[i] | clin.m[i,j] > ref.val$ref_high_M[i]
    }
  }
}
```

fPTrigly has age-dependent ref values and has to be fixed

```{r}

for(j in 1:ncol(clin.m)){
  if(clin.data$Sex[clin.data$ID==colnames(clin.m)[j]]=="F") {
    if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>20 &
       clin.data$Age[clin.data$ID==colnames(clin.m)[j]]<=30) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 1.5
    } else if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>30 &
              clin.data$Age[clin.data$ID==colnames(clin.m)[j]]<=50) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 1.7
    } else if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>50) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 2
    }
  } else if(clin.data$Sex[clin.data$ID==colnames(clin.m)[j]]=="M") {
    if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>20 &
       clin.data$Age[clin.data$ID==colnames(clin.m)[j]]<=30) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 1.7
    } else if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>30 &
              clin.data$Age[clin.data$ID==colnames(clin.m)[j]]<=50) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 2
    } else if(clin.data$Age[clin.data$ID==colnames(clin.m)[j]]>50) {
      ref.out["fPTrigly",j] <- clin.m["fPTrigly",j] > 2
    }
  }
}

```

The GEE estimates are consistent if the values are missing completely at random. However, the 5th visit has more missing values than the other. To stay on the safe side, I will remove the IID with incomplete longitudinal data. 

```{r}
for(i in unique(clin.data.long$IID)){
  if(length(unique(clin.data.long$Visit.number[clin.data.long$IID==i]))!=length(unique(clin.data.long$Visit.number))) {
    clin.data.long <- clin.data.long[clin.data.long$IID!=i,]
  }
}
```

Add "Out of range (OOR)" and "Out of range at baseline (OORB)" variables. In this case baseline will be visit 2.

```{r}
clin.data.long$OOR <- NA
clin.data.long$OORB <- NA
for(i in 1:nrow(ref.out)) {
  for(j in 1:ncol(ref.out)) {
    var.i <- rownames(ref.out)[i]
    ID.j <- colnames(ref.out)[j]
    IID.j <- clin.data$IID[ clin.data$ID %in% ID.j]
    ID.j.bas <- clin.data$ID[ clin.data$IID %in% IID.j & clin.data$Visit.number==2]
    clin.data.long$OOR[clin.data.long$variable==var.i & clin.data.long$ID==ID.j] <- ref.out[i,j]
    clin.data.long$OORB[clin.data.long$variable==var.i & clin.data.long$ID==ID.j] <- ref.out[i, colnames(ref.out) %in% ID.j.bas]
  }
}
```

## Fit GEE
I will model each variable as a function of `Sex`, `Age` and `Visit.number`. I will extract estimates for the whole polutation if there are no OOR at baseline (`OORB`), or if less than 5 individuals.

```{r}
list.gee <- list()
var <- unique(clin.data.long$variable)
for(i in 1:length(var)){
  
  # if no OOR at baseline or all OOR is NA
  if(all(!clin.data.long$OORB[clin.data.long$variable==var[i]]) |
     all(is.na(clin.data.long$OOR[clin.data.long$variable==var[i]]))) {
    
    data.tmp <- clin.data.long[clin.data.long$variable==var[i] & !is.na(clin.data.long$value),]
    data.tmp <- droplevels(data.tmp)
    data.tmp <- as.data.frame(data.tmp)
    
    list.gee[[i]]<- geeglm(value ~ Sex + Age + I(Visit.number-2),
                           id = as.factor(IID) , waves = Visit.number,
                           family = "gaussian", corstr = "exchangeable",
                           std.err = "san.se",
                           data=data.tmp)
    
  } else {
    
    data.tmp <- clin.data.long[clin.data.long$variable==var[i] & !is.na(clin.data.long$value) & !is.na(clin.data.long$OORB),]
    data.tmp <- droplevels(data.tmp)
    data.tmp <- as.data.frame(data.tmp)
    
    # if OORB=T for >= 5 IID
    if(length(unique(data.tmp$IID[data.tmp$OORB]))>=5) {
      list.gee[[i]]<- geeglm(value ~ Sex + Age + I(Visit.number-2)*OORB,
                           id = as.factor(IID) , waves = Visit.number,
                           family = "gaussian", corstr = "exchangeable",
                           std.err = "san.se",
                           data=data.tmp)
    } else {
      list.gee[[i]]<- geeglm(value ~ Sex + Age + I(Visit.number-2),
                           id = as.factor(IID) , waves = Visit.number,
                           family = "gaussian", corstr = "exchangeable",
                           std.err = "san.se",
                           data=data.tmp)
    }
  }
  
  rm(data.tmp)
}

names(list.gee) <-  var

summary.gee <- list()
for(i in 1:length(list.gee)) {
  summary.gee[[i]] <- summary(list.gee[[i]])
}
names(summary.gee) <- var

```

Test the marginal effect of `Visit.number`, as  a linear combination of `Visit.number` and `OORB`, if the interaction term was included.

```{r}

get.eff <- function(x) {
  if(length(coefficients(x))==4) {
    res <- esticon(x, c(0,0,0,1))
    type <- "All"
  } else {
    res <- esticon(x, c(0,0,0,1,0,1))
    type <- "OORB"
  }
  
  n <- length(unique(x$data$IID[x$data$OORB]))
  
  return(c(res,type=type, n_OORB=n))
}

effect.m <- sapply(list.gee, get.eff)
effect.df <- as.data.frame(t(effect.m))
effect.df$FDR <- p.adjust(effect.df[,"Pr(>|X^2|)"], method = "fdr")

DT::datatable(effect.df, options = list(scrollX=T))
```


For all the variables, print prediciton with bootstrapped SE. I will predict the effect for M and F and keeping `Age` constant and equal to the population mean.

```{r}

#sel.var <- na.omit(unique(clin.data.long$variable[clin.data.long$OORB]))
# sel.var <- c("ApoB_ApoA1ratio", "BMI", "DiastolicBP_mean", "fPGluk", "fPKol",
#              "fPKolHDL", "fpKolLDL",  "fPTrigly", "fSLipoA1", "fSLipoB", 
#              "hsCRP", "KolHDLrat", "PGT", "PULSE_mean", "SD25",
#              "STSH", "SystolicBP_mean")

new.data <- data.frame(Sex=c(rep("F", 8), rep("M", 8)),
                       Age=rep(mean(clin.data$Age), 16),
                       Visit.number=rep((2:5),4),
                       OORB=rep(c(rep(TRUE, 4), rep(FALSE, 4)),2))

bs.gee <- function(data, formula, newdata, indices) {
  d <- data[indices,]
  fit <- geeglm(formula,  id=as.factor(IID) , waves = Visit.number,
                family = "gaussian", corstr = "exchangeable",
                std.err = "san.se", data=d)
  predict <- predict(fit, newdata, type="response")
  return(predict)
}

list.gee.p <- vector("list",length(list.gee))
names(list.gee.p) <- names(list.gee)

for(var in names(list.gee)){

results <- boot(data = list.gee[[var]]$data,
                formula = list.gee[[var]]$formula,
                statistic=bs.gee, R=1000, newdata=new.data,
                parallel = "multicore",  ncpus = 36)


new.data$predict <- predict(list.gee[[var]], newdata=new.data, type="response")
new.data$lwr <- NA
new.data$upr <- NA
new.data$ref_low <- NA
new.data$ref_high <- NA

# get 95% confidence intervals and reference values
for(i in 1:nrow(new.data)){ try({
  new.data$lwr[i] <- boot.ci(results, type="bca", index=i)$bca[1,4]
  new.data$upr[i] <- boot.ci(results, type="bca", index=i)$bca[1,5]
  new.data$ref_low[i] <- ifelse(new.data$Sex[i]=="F",
                                ref.val$ref_low_F[match(var, ref.val$feature.ID)],
                                ref.val$ref_low_M[match(var, ref.val$feature.ID)])
  new.data$ref_high[i] <- ifelse(new.data$Sex[i]=="F",
                                 ref.val$ref_high_F[match(var, ref.val$feature.ID)],
                                 ref.val$ref_high_M[match(var, ref.val$feature.ID)])
})
  
}


list.gee.p[[var]] <- ggplot(new.data, aes(x=Visit.number, y=predict, group=factor(OORB, levels=c(TRUE, FALSE))))+
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=factor(OORB, levels=c(TRUE, FALSE))), alpha=0.2)+
  geom_line(aes(col=factor(OORB, levels=c(TRUE, FALSE))), lwd=2)+
  geom_hline(aes( yintercept= ref_low), lty=2) +
  geom_hline(aes(  yintercept= ref_high), lty=3) +
  facet_wrap(~Sex)+
  ggtitle(var)+
  scale_color_discrete(name="OORB")+
  scale_fill_discrete(name="OORB")+
  theme_bw()
print(list.gee.p[[var]])

}


```

## Fit Linear Mixed Model

```{r}
list.lmm <- list()
var <- unique(clin.data.long$variable)
for(i in 1:length(var)){
  
  # if no OOR at baseline or all OOR is NA
  if(all(!clin.data.long$OORB[clin.data.long$variable==var[i]]) |
     all(is.na(clin.data.long$OOR[clin.data.long$variable==var[i]]))) {
    
    data.tmp <- clin.data.long[clin.data.long$variable==var[i] & !is.na(clin.data.long$value),]
    data.tmp <- droplevels(data.tmp)
    data.tmp <- as.data.frame(data.tmp)
    
    list.lmm[[i]]<- lmer(value ~ Sex + Age + I(Visit.number-2) + (1 | IID),
                         data = data.tmp)
    
  } else {
    
    data.tmp <- clin.data.long[clin.data.long$variable==var[i] & !is.na(clin.data.long$value) & !is.na(clin.data.long$OORB),]
    data.tmp <- droplevels(data.tmp)
    data.tmp <- as.data.frame(data.tmp)
    
    # if OORB=T for >= 3 IID
    if(length(unique(data.tmp$IID[data.tmp$OORB]))>=5) {
      list.lmm[[i]] <- lmer(value ~ Sex + Age + I(Visit.number-2)*OORB + (1 | IID),
                          data = data.tmp)
    } else {
      list.lmm[[i]] <- lmer(value ~ Sex + Age + I(Visit.number-2) + (1 | IID),
                         data = data.tmp)
    }
  }
  
  rm(data.tmp)
}

names(list.lmm) <-  var



summary.lmm <- list()
for(i in 1:length(list.lmm)) {
  summary.lmm[[i]] <- summary(list.lmm[[i]])
}
names(summary.lmm) <- var

anova.lmm <- list()
for(i in 1:length(list.lmm)) {
  anova.lmm[[i]] <- Anova(list.lmm[[i]], test="F")
}
names(anova.lmm) <- var

```

Test the marginal effect of `Visit.number`, as  a linear combination of `Visit.number` and `OORB`, if the interaction term was included.

```{r}
get.eff <- function(x) {
  if(ncol(coefficients(x)$IID)==4) {
    res <- esticon(x, c(0,0,0,1))
    type <- "All"
  } else {
    res <- esticon(x, c(0,0,0,1,0,1))
    type <- "OORB"
  }
  
  n <- length(unique(x@frame$IID[x@frame$OORB]))
  
  return(c(res,type=type, n_OORB=n))
}

effect.m.lmm <- sapply(list.lmm, get.eff)
effect.df.lmm <- as.data.frame(t(effect.m.lmm))
effect.df.lmm$FDR <- p.adjust(effect.df.lmm[,"Pr(>|X^2|)"], method = "fdr")

DT::datatable(effect.df.lmm, options = list(scrollX=T))
```

Save

```{r}
save.image("/home/francesco.marabita/Projects/DHR/out/DHR_biometrics_GEE_v04.RData")
```

