---
title: "DHR clinical variables v01"
output:
  html_document:
    fig_width: 7
    fig_height: 7
    theme: cerulean
    toc: yes
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r}
library(readxl)
require(missMDA)
require(FactoMineR)
require(factoextra)
require(tidyverse)
```


## Read the data


```{r}
clin.data <- read_excel("/home/francesco.marabita/Projects/DHR/data/Clinical/All_visits_20180201.xlsx", 1, na = "NA")
```

Data must be fixed in the right format. Change the date to the correct format first (Ignore time zone. All times should be Finnish time). Change sex to factor.

```{r}
clin.data %>% mutate_at("dateofbirth", function(x) as.Date(x, format = "%Y-%m-%d")) %>%
  mutate_at("Visit", function(x) as.Date(x, format = "%d.%m.%Y")) %>%
  mutate(VisitDateTime = paste0(Visit, " ", Visittime)) %>%
  mutate_at("VisitDateTime", function(x) as.POSIXct(x, format = "%Y-%m-%d %H:%M", tz="Europe/Helsinki")) %>%
  mutate_at("Sex", as.factor) %>%
  mutate_at("Sex", function(x) recode_factor(x,"1"="M", "2"="F")) -> clin.data

```

Visit 3 and visit 5 are provided in a separate file.

```{r}
vis3 <- read_excel("/home/francesco.marabita/Projects/DHR/data/Clinical/Visit3_insulin.xlsx", 1, na = "NA")
vis5 <- read_excel("/home/francesco.marabita/Projects/DHR/data/Clinical/Visit5_insulin.xlsx", 1, na = "NA")
vis3.5 <- rbind(vis3, vis5)

tmp <- clin.data %>% 
  filter( VISITnumber==3 |  VISITnumber==5) %>%
  select(-SInsu) %>%
  left_join(vis3.5, by=c("visit_id"="subject"))

clin.data %<>%
  filter( VISITnumber<3 |  VISITnumber==4) %>%
  bind_rows(tmp)

rm(tmp)
```

## Factor Analysis of Mixed Data (FAMD)

I try Factor Analysis of Mixed data to visualize the data. I remove the ID variables and select only the quantitative and the qualitative variables. Select only mean values for blood pressure.

```{r}


to.FAMD <- clin.data %>% select(-ID,-visit_id, -VISITnumber, -Visittime) %>%
  select_if(function(x) is.character(x)|is.numeric(x)|is.factor(x)) %>%
  select(-matches("_[1-3]$")) %>%
  mutate_if(is.character, as.factor) %>% as.data.frame(.)
rownames(to.FAMD) <- paste0(clin.data$ID,"_", clin.data$VISITnumber)

#ncp <- estim_ncpFAMD(to.FAMD, pNA = 0.01, ncp.max = 5)

to.FAMD.imp <- imputeFAMD(to.FAMD, ncp = 5)

res.FAMD <- FAMD(to.FAMD, ncp = 15, tab.disj = to.FAMD.imp$tab.disj, graph = F)
```

```{r results="asis"}
DT::datatable(as.data.frame(res.FAMD$eig),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.FAMD$var$coord),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.FAMD$var$contrib),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.FAMD$var$cos2),
              options = list(autoWidth = TRUE, scrollX=TRUE))


```
### Visualize the results

Percentage ov variance explained and variable contribution

```{r}
fviz_screeplot(res.FAMD, ncp=15)

for(i in 1:15){
  print(fviz_contrib(res.FAMD, choice = "var", axes=i))
}

```

Relation between variables

```{r}
fviz_famd_var(res.FAMD, "quanti.var", col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = F)

plot(res.FAMD,choix = "quanti")

fviz_famd_var(res.FAMD, "quali.var", col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
plot(res.FAMD,choix = "quali")
```

Plot the individuals
```{r }
fviz_famd_ind(res.FAMD)
plot(res.FAMD, choix = "ind",lab.var = T)

for(i in 1:15){
  print(fviz_contrib(res.FAMD, choice = "ind", axes=i))
}
```

I will color the individuals based on all the variables.

```{r}
require(plotly)

toPlot <- cbind(to.FAMD.imp$completeObs, res.FAMD$ind$coord)
toPlot$label <- rownames(toPlot)

p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.2))+
  geom_point(aes(text=label,col=Sex))
ggplotly(p)

```


```{r fig.height=21, fig.width=7}
require(gridExtra)

p.l = list()
p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.2))
for(i in 1:(ncol(toPlot)-16)) {
  p.l[[i]] = p+geom_point(aes_string(col=colnames(toPlot)[i]))+
    theme(legend.position="bottom")
}
do.call(grid.arrange,c(p.l,ncol=4, nrow=10))

```

Same but with Dim2 - Dim3

```{r}

p <- ggplot(toPlot, aes(x=Dim.3, y=Dim.4))+
  geom_point(aes(text=label,col=Sex))
ggplotly(p)
```

```{r fig.height=21, fig.width=7}

p.l = list()
p <- ggplot(toPlot, aes(x=Dim.3, y=Dim.4))
for(i in 1:(ncol(toPlot)-16)) {
  p.l[[i]] = p+geom_point(aes_string(col=colnames(toPlot)[i]))+
    theme(legend.position="bottom")
}
do.call(grid.arrange,c(p.l,ncol=4, nrow=10))

```

I repeat the same analysis, but this time I do not use the last visit for factor analysis. The samples from visit 5 will be only projected into the new space
```{r}

ind.sup <- which(clin.data$VISITnumber==5)
res.FAMD.bas <- FAMD(to.FAMD.imp$completeObs, ncp = 15,
                     ind.sup = ind.sup)
summary(res.FAMD.bas)

```

## PCA

Impute missing data first and then perform PCA (with base or FactoMineR packages)

```{r}
to.PCA <- clin.data %>% select(-ID,-visit_id, -VISITnumber, -Visittime) %>%
  select_if(is.numeric) %>%
  select(-matches("_[1-3]$")) %>%
  as.data.frame()

rownames(to.PCA) <- paste0(clin.data$ID,"_", clin.data$VISITnumber)

estim_ncpPCA(to.PCA, ncp.max = 10, method.cv = "Kfold", scale=T)

to.PCA.imp <- imputePCA(to.PCA, ncp = 5, scale = T)
```

With base package

```{r}
res.PCA.base <- prcomp(to.PCA.imp$completeObs,scale. = T, center = T)

biplot(res.PCA.base, cex = c(0.5,1))
```

With FactoMineR

```{r}
res.PCA<- PCA(to.PCA.imp$completeObs, scale.unit = T, graph = F, ncp = 15)
```

```{r results="asis"}
DT::datatable(as.data.frame(res.PCA$eig),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.PCA$var$coord),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.PCA$var$contrib),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(res.PCA$var$cos2),
              options = list(autoWidth = TRUE, scrollX=TRUE))


```

### Visualize the results

```{r}
fviz_screeplot(res.PCA,  ncp = 15)

for(i in 1:15){
  print(fviz_contrib(res.PCA, choice = "var", axes=i))
}

```
```{r}
fviz_pca_biplot(res.PCA, col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = F,label = "var")
```

Relation between variables

```{r}
fviz_pca_var(res.PCA, col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = F)

```

Plot the individuals

```{r }
fviz_pca_ind(res.PCA)

for(i in 1:15){
  print(fviz_contrib(res.PCA, choice = "ind", axes=i))
}
```

```{r}
require(plotly)

toPlot <- data.frame(to.PCA.imp$completeObs, Sex=clin.data$Sex, res.PCA$ind$coord)
toPlot$label <- rownames(toPlot)
toPlot$Sex <- relevel(toPlot$Sex,ref = "F")

p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.2))+
  geom_point(aes(text=label,col=Sex))
ggplotly(p)

```

```{r}
p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.3))+
  geom_point(aes(text=label,col=Sex))
ggplotly(p)

```

```{r}
p <- ggplot(toPlot, aes(x=Dim.3, y=Dim.4))+
  geom_point(aes(text=label,col=Sex))
ggplotly(p)

```

```{r fig.height=21, fig.width=7}

p.l = list()
p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.2))
for(i in 1:(ncol(toPlot)-16)) {
  p.l[[i]] = p+geom_point(aes_string(col=colnames(toPlot)[i]))+
    theme(legend.position="bottom")
}
do.call(grid.arrange,c(p.l,ncol=4, nrow=10))

```

```{r fig.height=21, fig.width=7}

p.l = list()
p <- ggplot(toPlot, aes(x=Dim.1, y=Dim.3))
for(i in 1:(ncol(toPlot)-16)) {
  p.l[[i]] = p+geom_point(aes_string(col=colnames(toPlot)[i]))+
    theme(legend.position="bottom")
}
do.call(grid.arrange,c(p.l,ncol=4, nrow=10))

```

```{r fig.height=21, fig.width=7}

p.l = list()
p <- ggplot(toPlot, aes(x=Dim.3, y=Dim.4))
for(i in 1:(ncol(toPlot)-16)) {
  p.l[[i]] = p+geom_point(aes_string(col=colnames(toPlot)[i]))+
    theme(legend.position="bottom")
}
do.call(grid.arrange,c(p.l,ncol=4, nrow=10))

```

