---
title: "DHR MOFA2 prepare the data v02"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 9
    fig_height: 7
    theme: cerulean
    toc: yes
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

Require all the libraries first
```{r message=F}
require(MultiAssayExperiment)
require(SummarizedExperiment)
require(MOFA2)
require(RColorBrewer)
require(tidyverse)
```

# Load the data

```{r}
load("/home/francesco.marabita/Projects/DHR/data/MultiAssayExp/out/DHR_MAexp_v05.RData")
outDir <-  "/home/francesco.marabita/Projects/DHR/out/MOFA2_v02"
dir.create(outDir)
```

# Create MOFA object

For this analyis I will consider the Autoantibody measurements as a binary feature. Therefore I will consider the "Binary" assay.

```{r}
tmp <- assays(MAexp[["autoab"]])$Binary
tmp <- tmp[rownames(tmp) %in% rownames(MAexp.clean[["autoab"]]),]
if(all(rownames(tmp)==rownames(MAexp.clean[["autoab"]]))) {
  assays(MAexp.clean[["autoab"]]) <- list("Binary" = tmp)
} else {message("Rownames are different")}

```

I will remove the feature with zero variance and the clinical data.

```{r}
MAexp.clean <- MAexp.clean[,,!names(MAexp.clean) %in% "clinical"]

keep.col <- intersect(intersect(intersect(colnames(assay(MAexp.clean[["GCMS"]])), colnames(assay(MAexp.clean[["autoab"]]))), colnames(assay(MAexp.clean[["LCMS.pos"]]))), colnames(assay(MAexp.clean[["microb.16S"]])))

for (i in 1:length(experiments(MAexp.clean))){
  experiments(MAexp.clean)[[i]] <- experiments(MAexp.clean)[[i]][,colnames(MAexp.clean[[i]]) %in% keep.col]
  experiments(MAexp.clean)[[i]] <- experiments(MAexp.clean)[[i]][apply(assay(MAexp.clean[[i]]), 1, function(x) var(x,na.rm = T) != 0), ]
}
```

# Create MOFA object

A data frame with columns "sample","feature","view","group","value" needs to be created and saved as text files.

```{r}
assay(MAexp.clean[["GCMS"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".GCMS")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="GCMS",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> gcms.df

assay(MAexp.clean[["LCMS.neg"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".LCMS.neg")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="LCMS.neg",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> lcms.neg.df

assay(MAexp.clean[["LCMS.pos"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".LCMS.pos")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="LCMS.pos",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> lcms.pos.df

assay(MAexp.clean[["autoab"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".autoab")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="Autoantibodies",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> autoab.df

assay(MAexp.clean[["prot.1"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.1")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.1.df

assay(MAexp.clean[["prot.2"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.2")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.2.df

assay(MAexp.clean[["prot.3"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.3")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.3.df

assay(MAexp.clean[["prot.4"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.4")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.4.df

assay(MAexp.clean[["prot.5"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.5")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.5.df

assay(MAexp.clean[["prot.6"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".prot.6")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="PEA",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> prot.6.df

assay(MAexp.clean[["microb.16S"]]) %>%
  as.data.frame() %>% 
  rownames_to_column("feature.ID") %>%
  mutate(feature = paste0(feature.ID, ".microb.16S")) %>% 
  gather(key="sample", value = "value", -c(feature, feature.ID)) %>% 
  mutate(view="microb.16S",
         group="DHR") %>% 
  select(sample, feature, view, group, value) -> microb.16S.df

mofa.input.df <- rbind(gcms.df,
                       lcms.neg.df,
                       lcms.pos.df,
                       autoab.df,
                       prot.1.df, prot.2.df, prot.3.df, prot.4.df, prot.5.df, prot.6.df,
                       microb.16S.df)

```

Create annotation data frame.

```{r}
# metabolite category
load("/home/francesco.marabita/Projects/DHR/ext_data/hmdb_selected_fields.RData")
hmdb.df$category <- ifelse(hmdb.df$sub.class=="", hmdb.df$class, hmdb.df$sub.class)
tmp <- annot.df[, c("featureID", "HMDB.ID.1")]
tmp <- tmp[!is.na(tmp$HMDB.ID.1),]
tmp$HMDB.ID.new <- sub("HMDB", "HMDB00", tmp$HMDB.ID.1)
tmp$metab.category <- hmdb.df$category[match(tmp$HMDB.ID.new, hmdb.df$accession)]


mofa.annot <- data.frame(feature = unique(mofa.input.df$feature), stringsAsFactors = F)
mofa.annot$feature.ID <- mofa.annot$feature

mofa.annot$feature.ID <- sub("\\.GCMS$","", mofa.annot$feature.ID)
mofa.annot$feature.ID <- sub("\\.LCMS\\.neg$","", mofa.annot$feature.ID)
mofa.annot$feature.ID <- sub("\\.LCMS\\.pos$","", mofa.annot$feature.ID)
mofa.annot$feature.ID <- sub("\\.autoab$","", mofa.annot$feature.ID)
mofa.annot$feature.ID <- sub("\\.prot\\..+$","", mofa.annot$feature.ID)
mofa.annot$feature.ID <- sub("\\.microb\\.16S$","", mofa.annot$feature.ID)

mofa.annot %>% 
  left_join(annot.df, by=c("feature.ID"="featureID")) %>% 
  distinct() %>%
  mutate(metab.category = tmp$metab.category[match(feature.ID, tmp$featureID)]) -> mofa.annot
```

Create MOFA data

```{r}
write.table(mofa.input.df, file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_input.txt", sep="\t", quote=F, row.names = F)
write.table(mofa.annot, file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_annotation.txt", sep="\t", quote=F, row.names = F)

mofa.data <- create_mofa(mofa.input.df)
mofa.data
plot_data_overview(mofa.data)
```
