---
title: "DHR process phenotypes for MOFA interpetation v03"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 7
    fig_height: 7
    theme: cerulean
    toc: yes
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r}
require(openxlsx)
require(lubridate)
require(tidyverse)

load("/home/francesco.marabita/Projects/DHR/data/Clinical/Clin_data_2018-10-09_newVarNames.RData")

outDir <-  "/home/francesco.marabita/Projects/DHR/out/pheno_preproc_v03"
dir.create(outDir)

```

# Questionnaire

Import the questionaire data and select the question related to:

* Family history
* Exercise and physical activity
* Functional capacity and health
* Mental health
* Diet
* Smoking
* Alcohol use
* Sleep habits

```{r}

v1 <- read.table("/home/francesco.marabita/Projects/DHR/data/Questionnaire/Code_headers/kyselylomake_1_20181120_145706.txt",
                 sep="\t", header=T)
v2 <- read.table("/home/francesco.marabita/Projects/DHR/data/Questionnaire/Code_headers/kyselylomake_2_20181120_150251.txt",
                 sep="\t", header=T)
v3 <- read.table("/home/francesco.marabita/Projects/DHR/data/Questionnaire/Code_headers/kyselylomake_3_20181120_150756.txt",
                 sep="\t", header=T)
v4 <- read.table("/home/francesco.marabita/Projects/DHR/data/Questionnaire/Code_headers/kyselylomake_4_20181120_151120.txt",
                 sep="\t", header=T)
v5 <- read.table("/home/francesco.marabita/Projects/DHR/data/Questionnaire/Code_headers/kyselylomake_5_20181122_103240.txt",
                 sep="\t", header=T)
```

The variables may have different IDs between visits. Import lookup table with selected questions.

```{r}
sel.conv <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Questionnaire/selected_questions_conversion.xlsx")
sel.conv$ID1 <- gsub("-", ".", sel.conv$ID1)
sel.conv$ID2 <- gsub("-", ".", sel.conv$ID2)
sel.conv$ID3 <- gsub("-", ".", sel.conv$ID3)
sel.conv$ID4 <- gsub("-", ".", sel.conv$ID4)
sel.conv$ID5 <- gsub("-", ".", sel.conv$ID5)
```

Select the questions and make long format. 

```{r}
v1 <- v1[,colnames(v1) %in% c("subject", sel.conv$ID1)]
v2 <- v2[,colnames(v2) %in% c("subject", sel.conv$ID2)]
v3 <- v3[,colnames(v3) %in% c("subject", sel.conv$ID3)]
v4 <- v4[,colnames(v4) %in% c("subject", sel.conv$ID4)]
v5 <- v5[,colnames(v5) %in% c("subject", sel.conv$ID5)]

v1.l <- gather(v1, key = "question_ID1", value="value", -subject)
# question ID changed during visit 2:fix it

v1.l$question_ID <- sel.conv$ID2[match(v1.l$question_ID1, sel.conv$ID1)]
v1.l <- v1.l[,c("subject", "question_ID", "value")]
v2.l <- gather(v2, key = "question_ID", value="value", -subject)
v3.l <- gather(v3, key = "question_ID", value="value", -subject)
v4.l <- gather(v4, key = "question_ID", value="value", -subject)
v5.l <- gather(v5, key = "question_ID", value="value", -subject)

rbind(v1.l, v2.l, v3.l, v4.l, v5.l) %>% 
  rename(ID = subject) -> v1to5.l
```

Add type.

```{r}
v1to5.l$Type <- sel.conv$Type[match(v1to5.l$question_ID, sel.conv$ID2)]
v1to5.l$Units <- sel.conv$Units[match(v1to5.l$question_ID, sel.conv$ID2)]
```


Remove the text for each answer and keep the label only.

```{r}
v1to5.noText <- v1to5.l
v1to5.noText$value <- gsub(" \\(.*)$", "", v1to5.noText$value)
```


# Healthy Food Intake Index (HFII)

```{r}
hfii<- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/Healthy_Food_Intake_Index_components.xlsx", sheet = 1)
hfii.l <- gather(hfii, key="key", value="value", -IID)
hfii.l$name <- paste0(hfii.l$IID,"_", substr(hfii.l$key,2,2))

lookup.id <- clin.data[,c("IID","ID","Visit.number")]
lookup.id$name <- paste0(lookup.id$IID, "_", lookup.id$Visit.number)

hfii.l$ID <- lookup.id$ID[match(hfii.l$name, lookup.id$name)]
#hfii.l <- hfii.l[, c("IID","ID","name", "value")]
hfii.l <- hfii.l[!is.na(hfii.l$ID),]
```

# Menstrual cycle

```{r}
m1 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/menstrual_cycle.xlsx", sheet = "v1", detectDates = T )
m2 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/menstrual_cycle.xlsx", sheet = "v2", detectDates = T )
m3 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/menstrual_cycle.xlsx", sheet = "v3", detectDates = T )
m4 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/menstrual_cycle.xlsx", sheet = "v4", detectDates = T )
m5 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/menstrual_cycle.xlsx", sheet = "v5", detectDates = T )

m1 <- m1[, c("ID", "start_date_menstruation" )]
m1$note_menstruation <-  ""
m2 <- m2[, c("ID", "start_date_menstruation", "note_menstruation" )]
m3 <- m3[, c("ID", "start_date_menstruation", "note_menstruation" )]
m4 <- m4[, c("ID", "start_date_menstruation", "note_menstruation" )]
m5 <- m5[, c("ID", "start_date_menstruation", "note_menstruation" )]

m1to5 <- rbind(m1,m2,m3,m4,m5)
m1to5$IID <- lookup.id$IID[match(m1to5$ID, lookup.id$ID)]
```


# Saliva cortisol

## $$AUC_G$$ and $$AUC_I$$

$$AUC_G = \sum_{i=1}^{n-1} \frac{(m_{i+1} + m_i) \cdot t_i} {2}$$
$$AUC_I = AUC_G - m_1 \cdot \sum_{i=1}^{n-1}t_i$$

$$`t_i`$$ is the time distance between measurements $$m_{i+1}$$ and $$m_i$$.

```{r}
s1to5 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Saliva_cortisol_processed_2020-02-05.txt",
                    header = T, sep = "\t", stringsAsFactors = F)

s1to5 %>% 
  rename(ID=subject) %>% 
  select(ID,collection_date,procesing_date,
         C1.1_arvo,C1.1_protocol,C1.1_tstamp,
         C2.1_arvo,C2.1_protocol,C2.1_tstamp,
         C3.1_arvo,C3.1_protocol,C3.1_tstamp,
         C4.1_arvo,C4.1_protocol,C4.1_tstamp,
         C.any.flag.count, C.interval.flag.count, C.interval.times,
         comment) %>% 
  filter(C.any.flag.count==0) -> s1to5
  
# convert to times
s1to5$C1.1_datetime <- as.POSIXct(strptime(paste0(s1to5$collection_date," ",s1to5$C1.1_tstamp), format="%Y-%m-%d %H:%M"))
s1to5$C2.1_datetime <- as.POSIXct(strptime(paste0(s1to5$collection_date," ",s1to5$C2.1_tstamp), format="%Y-%m-%d %H:%M"))
s1to5$C3.1_datetime <- as.POSIXct(strptime(paste0(s1to5$collection_date," ",s1to5$C3.1_tstamp), format="%Y-%m-%d %H:%M"))
s1to5$C4.1_datetime <- as.POSIXct(strptime(paste0(s1to5$collection_date," ",s1to5$C4.1_tstamp), format="%Y-%m-%d %H:%M"))

# if C4 is after midnight, add one day
s1to5$C4.1_datetime <- if_else((s1to5$C4.1_datetime - s1to5$C1.1_datetime) <= 0,
                              s1to5$C4.1_datetime +  hm("24:00"),
                              s1to5$C4.1_datetime)


# calculate AUC_G and AUC_I
s.score <- data.frame(ID=s1to5$ID, AUC.g=NA, AUC.i=NA, C.1=NA, C.4=NA, C.p=NA, C.D=NA)
for(i in 1:nrow(s1to5)){
   if(any(is.na(c(s1to5$C1.1_arvo[i], s1to5$C2.1_arvo[i], s1to5$C3.1_arvo[i], s1to5$C4.1_arvo[i])))) {
    s.score$AUC.g[i] <- NA
    s.score$AUC.i[i] <- NA
  } else {
    s.score$AUC.g[i] <- sum(
      ((s1to5$C2.1_arvo[i] + s1to5$C1.1_arvo[i])*(as.numeric(difftime(s1to5$C2.1_datetime[i], s1to5$C1.1_datetime[i], units = "mins"))))/2, 
      ((s1to5$C3.1_arvo[i] + s1to5$C2.1_arvo[i])*(as.numeric(difftime(s1to5$C3.1_datetime[i], s1to5$C2.1_datetime[i], units = "mins"))))/2,
      ((s1to5$C4.1_arvo[i] + s1to5$C3.1_arvo[i])*(as.numeric(difftime(s1to5$C4.1_datetime[i], s1to5$C3.1_datetime[i], units = "mins"))))/2
    )
    s.score$AUC.i[i] <- s.score$AUC.g[i] - s1to5$C1.1_arvo[i]*(as.numeric(difftime(s1to5$C4.1_datetime[i], s1to5$C3.1_datetime[i], units = "mins")))
  }
}

```

## Awakening, evening, peak and Delta (peak - ground) levels 

```{r}
s.score$C.1 <- s1to5$C1.1_arvo
s.score$C.4 <- s1to5$C4.1_arvo
s.score$C.p <- pmax(s1to5$C1.1_arvo, s1to5$C2.1_arvo, s1to5$C3.1_arvo, s1to5$C4.1_arvo, na.rm = T)
s.score$C.D <- pmax(s1to5$C1.1_arvo, s1to5$C2.1_arvo, s1to5$C3.1_arvo, s1to5$C4.1_arvo, na.rm = T) -
   pmin(s1to5$C1.1_arvo, s1to5$C2.1_arvo, s1to5$C3.1_arvo, s1to5$C4.1_arvo, na.rm = T)

```


# Apnea
```{r}
a1to5 <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Other/Selfreported sleep apnea score_DHR_OSA_scores_Q1-Q5.xlsx",
                   sheet = 1)
a1to5 <- a1to5[,3:ncol(a1to5)]
a.score <- as.data.frame(mapply(c, a1to5[,1:2],a1to5[,3:4], a1to5[,5:6], a1to5[,7:8], a1to5[,9:10]))
colnames(a.score) <- c("ID", "apnea_score")
```

# Fitness

```{r}
f1 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Fitness/kuntotesti_1_20181120_152601.txt",
                 sep="\t", header=T)
f2 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Fitness/kuntotesti_2_20181120_152622.txt",
                 sep="\t", header=T)
f3 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Fitness/kuntotesti_3_20181120_152652.txt",
                 sep="\t", header=T)
f4 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Fitness/kuntotesti_4_20181120_152739.txt",
                 sep="\t", header=T)
f5 <- read.table("/home/francesco.marabita/Projects/DHR/data/Other/Fitness/kuntotesti_5_20181120_152815.txt",
                 sep="\t", header=T)
f1to5 <- rbind(f1,f2,f3,f4,f5)
colnames(f1to5) <- c("ID", "state", "timestamp", "central_body", "mobility", "balance", "lower_body", "upper_body")
f1to5 %>% 
  select(-state, -timestamp) %>% 
  gather(key="component", value = "value", -ID) -> f1to5.l

```


```{r}
save(sel.conv, v1to5.l, v1to5.noText, hfii.l, m1to5, s.score, a.score, f1to5.l, file=paste0(outDir,"/DHR_pheno_preproc_v03.RData"))
```

