---
title: "DHR MOFA2 v02 downstream analysis"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 9
    fig_height: 7
    theme: cerulean
    toc: yes
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

Require all the libraries first
```{r message=F}
require(MultiAssayExperiment)
require(SummarizedExperiment)
require(MOFA2)
require(RColorBrewer)
require(tidyverse)
```

# Load the data

```{r}
load("/home/francesco.marabita/Projects/DHR/data/MultiAssayExp/out/DHR_MAexp_v05.RData")
mofa.annot <- read.table(file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_annotation.txt", sep="\t", header = T, stringsAsFactors = F, comment.char ="", quote = "")

outDir <-  "/home/francesco.marabita/Projects/DHR/out/MOFA2_v02"
```

Based on preliminary exploration, he models without scaling have a better ELBO value. I will load the trained models with with `iter=10000`, `convergence_mode = "medium",` and `dropR2=0.02`.

```{r}
files <- list.files(outDir, pattern = "^.*_train_[0-9]+\\.hdf5$", full.names = T)
mofa.train.l <- list()
for(i in 1:length(files)){
  mofa.train.l[[i]] <- load_model(files[i])
  print(sapply(strsplit(files[i], split = "/", fixed = T), "[", 8))
  names(mofa.train.l)[i] <- sapply(strsplit(files[i], split = "/", fixed = T), "[", 8)
}


elbo.df <- compare_elbo(mofa.train.l, log = F,return_data = T)
elbo.df$num.factors <- sapply(mofa.train.l, function(x) get_dimensions(x)["K"])

DT::datatable(elbo.df)

elbo.df[which.min(elbo.df$ELBO),]

compare_factors(mofa.train.l, show_colnames=F)

mofa.sel <- select_model(mofa.train.l)
plot_factor_cor(mofa.sel)
```


Add the metadata to the model

```{r}
sample_metadata <- mofa.sel@samples_metadata

MAexp[["clinical"]] %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  right_join(as.data.frame(pData)) -> tmp

sample_metadata <- merge(sample_metadata, tmp, by.x="sample", by.y="ID", all.x=T, sort=F)
sample_metadata$pheno_subtype <- fct_recode(sample_metadata$clust.mf.comb, Low = "F1", High = "F2", High="F3", High="F4", Low="M1", High="M2")


if(all(sample_metadata$sample==mofa.sel@samples_metadata$sample)){
  samples_metadata(mofa.sel) <- as.data.frame(sample_metadata)
} else { message("Sample order does not match")}

features_metadata <-  mofa.sel@features_metadata
features_metadata <- merge(features_metadata, mofa.annot, by="feature", all.x=T, sort=F)
if(all(features_metadata$feature==mofa.sel@features_metadata$feature)){
  features_metadata(mofa.sel) <- as.data.frame(features_metadata)
} else { message("Sample order does not match")}

```


# Analysis

## Variance explained

Check the variance epxlained.

```{r}
#r2 <- mofa.sel@cache$variance_explained
r2 <- get_variance_explained(mofa.sel)
# r2
df.expl.var <- as.data.frame(r2$r2_per_factor)
df.expl.var$factor <- rownames(df.expl.var)
DT::datatable(df.expl.var)

p <- plot_variance_explained(mofa.sel, cluster = F, plot_total = T)
p[[1]]
p[[2]]

```

## Inspect the factors

```{r}
p <- plot_factor(mofa.sel, factors = 1:mofa.sel@dimensions$K,  color_by = "IID")
p$layers[[1]]$aes_params$alpha <- 0.7
p + scale_fill_viridis_d() + theme(legend.position = "none")

p <- plot_factor(mofa.sel, factors = 1:mofa.sel@dimensions$K,  color_by = "Age")
p$layers[[1]]$aes_params$alpha <- 0.7
p + scale_fill_viridis_c()

p <- plot_factor(mofa.sel, factors = 1:mofa.sel@dimensions$K,  color_by = "Sex")
p$layers[[1]]$aes_params$alpha <- 0.7
p + scale_fill_manual("Sex", values=c("F"="#d7191c", "M"="#2c7bb6"))

p <- plot_factor(mofa.sel, factors = 1:mofa.sel@dimensions$K,  color_by = "BMI")
p$layers[[1]]$aes_params$alpha <- 0.7
p + scale_fill_viridis_c()

p <- plot_factor(mofa.sel, factors = 1:mofa.sel@dimensions$K,  color_by = "pheno_subtype")
p$layers[[1]]$aes_params$alpha <- 0.7
p + scale_fill_manual("Subtype", values=c("High"="#d7191c", "Low"="#2c7bb6"))
```


## Inspect the weights

Visualize the top weights

```{r}
for(i in 1:mofa.sel@dimensions$K){
  for(j in 1:length(views_names(mofa.sel))){
    p <- plot_top_weights(mofa.sel, factor = i, nfeatures = 30, view = views_names(mofa.sel)[j])
    p <- p + scale_x_discrete(labels = rev(mofa.annot$feature.name[match(p$data$feature, mofa.annot$feature)])) +
    ggtitle(views_names(mofa.sel)[j]) +
    theme(axis.text.y=element_text(size = rel(0.9)))
    print(p)
  }
}
```

```{r}
for(i in 1:mofa.sel@dimensions$K){
  for(j in 1:length(views_names(mofa.sel))){
    p <- plot_weights(mofa.sel, factor = i, nfeatures = 10, view = views_names(mofa.sel)[j], scale = T, abs=F) + 
      ggtitle(paste0(views_names(mofa.sel)[j], " - Factor ",i))

  p$layers[[2]]$data$feature <- mofa.annot$feature.name[match(p$layers[[2]]$data$feature, mofa.annot$feature)]
  print(p)
  }
}

```

Table of the weight for the top 50 features.

```{r}
get_weights(mofa.sel, as.data.frame = T, scale=F, abs=F) %>%
  mutate(fearure.ID=mofa.annot$feature.ID[match(feature, mofa.annot$feature)],
         feature.name=mofa.annot$feature.name[match(feature, mofa.annot$feature)]) -> f.weights

w.top50.l <- list()
for(i in 1:mofa.sel@dimensions$K){
  w.top50.l[[i]] <- arrange(f.weights[f.weights$factor==paste0("Factor",i),], desc(abs(value)))[1:50,]
  names(w.top50.l)[i] <- paste0("Factor",i)
}

htmltools::tagList(
  lapply(w.top50.l, function(x) DT::datatable(x, options =list(pageLength=10) ))
)

```


## Sample ordination

```{r}
set.seed(12345)
mofa.sel <- run_umap(mofa.sel, n_neighbors = 20)

plot_dimred(mofa.sel, method = "UMAP",  color_by = "IID", color_name = "IID") +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

plot_dimred(mofa.sel, method = "UMAP",  color_by = "Age", color_name = "Age") +
  scale_fill_viridis_c()

plot_dimred(mofa.sel, method = "UMAP",  color_by = "Sex", color_name = "Sex") +
  scale_fill_viridis_d()

plot_dimred(mofa.sel, method = "UMAP",  color_by = "BMI", color_name = "BMI") +
  scale_fill_viridis_c()

plot_dimred(mofa.sel, method = "UMAP",  color_by = "pheno_subtype", color_name = "Subtype") +
  scale_fill_manual("Subtype", values=c("High"="#d7191c", "Low"="#2c7bb6"))

```


## Inspect the single views and the loadings

```{r}

for(v in 1:length(views_names(mofa.sel))){
  
  plot_weights_heatmap(
    mofa.sel, 
    view = views_names(mofa.sel)[v],
    factors = 1:mofa.sel@dimensions$K,
    show_colnames = F,
    main=views_names(mofa.sel)[v]
  )
  
  
  p <- plot_top_weights(mofa.sel, view =  views_names(mofa.sel)[v],
                        factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[1],
                        nfeatures = 50)
  plot_data_heatmap(mofa.sel, view =  views_names(mofa.sel)[v],
                    factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[1],
                    features = 50, labels_row = mofa.annot$feature.name[match(p$data$feature, mofa.annot$feature)],
                    main = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[1],
                    scale=ifelse(v==1,"none","row"), color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdBu")))(100))
  
  p <- plot_top_weights(mofa.sel, view =  views_names(mofa.sel)[v],
                        factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[2],
                        nfeatures = 50)
  plot_data_heatmap(mofa.sel, view =  views_names(mofa.sel)[v],
                    factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[2],
                    features = 50, labels_row = mofa.annot$feature.name[match(p$data$feature, mofa.annot$feature)],
                    main = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[2],
                    scale=ifelse(v==1,"none","row"), color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdBu")))(100))
  
  p <- plot_top_weights(mofa.sel, view =  views_names(mofa.sel)[v],
                        factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[3],
                        nfeatures = 50)
  plot_data_heatmap(mofa.sel, view =  views_names(mofa.sel)[v],
                    factor = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[3],
                    features = 50, labels_row = mofa.annot$feature.name[match(p$data$feature, mofa.annot$feature)],
                    main = rownames(df.expl.var[order(df.expl.var[,v],decreasing = T),])[3],
                    scale=ifelse(v==1,"none","row"), color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdBu")))(100))
  
}
```

```{r}
# save(mofa.sel, file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_sel_model.RData")
# saveRDS(mofa.sel, file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_sel_model.rds")
# mofa.factors <- get_factors(mofa.sel)$DHR
# save(mofa.factors, file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02_factors.RData")
# save.image(file="~/Projects/DHR/out/MOFA2_v02/DHR_MOFA2_v02.RData")
```



