---
title: "DHR Proteomics v01"
output:
  html_document:
    fig_width: 9
    fig_height: 7
    theme: cerulean
    toc: yes
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

Require all the libraries
```{r}
require(SummarizedExperiment)
require(plotly)
require(geneplotter)
require(RColorBrewer)
require(openxlsx)
require(factoextra)
require(FactoMineR)
require(ICC)
require(tidyverse)
```

Read the data and store as SummarizedEzxperiment

```{r}
outDir <- "/home/francesco.marabita/Projects/DHR/data/Proteomics/out"
load("/home/francesco.marabita/Projects/DHR/data/Clinical/Clin_data_2018-02-13.RData")


prot.files <- list.files("/home/francesco.marabita/Projects/DHR/data/Proteomics", pattern ="^Proteomics_.*\\.xlsx$", full.names = T)

for(i in 1:length(prot.files)){
  # read data

  assayData_NaN <- read.xlsx(prot.files[i], sheet = "assayData_NaN", colNames =T, rowNames = T,
                             skipEmptyRows=T, na.strings = "NaN")
  assayData_NaN[assayData_NaN=="NaN"] <- NA
  rownames(assayData_NaN) <- make.names(rownames(assayData_NaN))
  

  assayData_LOD <- read.xlsx(prot.files[i], sheet = "assayData_LOD", colNames = T, rowNames = T,
                             skipEmptyRows=T, na.strings = "NaN")
  rownames(assayData_LOD) <- make.names(rownames(assayData_LOD))

  fData <- read.xlsx(prot.files[i], sheet = "fData", colNames =T, rowNames = F, skipEmptyRows=T)
  fData$orig.feature.ID <- fData$feature.ID
  fData$feature.ID <- make.names(fData$orig.feature.ID)
  rownames(fData) <- fData$feature.ID
  
  pData_ <- read.xlsx(prot.files[i], sheet = "pData", colNames =T, rowNames = F, skipEmptyRows=T)
  
  tmp <- as.data.frame(clin.data[, c("ID", "visit_id", "VISITnumber", "dateofbirth", "Age", "Visit", "Visittime", "Sex")])
  colnames(tmp) <- c("IID", "ID", "Visit.number", "Birth.date", "Age", "Visit.date", "Visit.time", "Sex")

  pData <- merge(pData_, tmp, by.x="Visit.ID", by.y="ID", all.x=T, sort=F)
  pData <- pData[match(pData_$Visit.ID, pData$Visit.ID),]
  colnames(pData)[1] <- "ID"
  rownames(pData) <- pData$ID
  
 
  gc()
  
  # create SummarizedExperiment
  if(all(rownames(assayData_NaN)==rownames(assayData_LOD)) &
     all(colnames(assayData_NaN)==colnames(assayData_LOD)) &
     all(rownames(assayData_NaN)==rownames(fData)) &
     all(colnames(assayData_NaN)==rownames(pData))){
    
    tmp <- SummarizedExperiment(assays = list(NPX_NaN=as.matrix(assayData_NaN), 
                                              NPX_LOD=as.matrix(assayData_LOD)),
                                rowData = DataFrame(fData),
                                colData = DataFrame(pData),
                                metadata = list(panel=sapply(strsplit(prot.files[i], "_"), "[",2)))
    
    tmp <- tmp[,tmp$IID %in% unique(clin.data$ID)]
    rowData(tmp)$n.LOD <- rowSums(is.na(assays(tmp)$NPX_NaN))
    assign(paste0("prot.se.",i), tmp)
    rm(tmp)
    
  } else { stop("Order is different") }
  
  message(paste0("Done with ", i, " of ", length(prot.files)))
}

```
Check overlap between proteins

```{r}
l.fID <- list()
l.uID <- list()
for(i in 1:length(prot.files)){
  l.fID[[i]] <- rowData(get(paste0("prot.se.",i)))$feature.ID
  l.uID[[i]] <- rowData(get(paste0("prot.se.",i)))$Uniprot.ID
  names(l.fID)[i] <-metadata(get(paste0("prot.se.",i)))$panel
  names(l.uID)[i] <-metadata(get(paste0("prot.se.",i)))$panel

}

crossprod(table(stack(l.fID)))
crossprod(table(stack(l.uID)))



```

# Quality control
In order to check for sample-to-sample relationship, outliers and batch effects, I consider the following:

* Density plot
* Boxpots
* PCA

## Density plot

```{r}
for(i in 1:length(prot.files)){
  multidensity(assays(get(paste0("prot.se.",i)))$NPX_LOD, legend = F, main=metadata(get(paste0("prot.se.",i)))$panel,
               xlab="NPX")
  multidensity(assays(get(paste0("prot.se.",i)))$NPX_NaN, legend = F, main=metadata(get(paste0("prot.se.",i)))$panel,
               xlab="NPX")
}
```

## Boxplots
```{r  fig.height=20}
# for(i in 1:length(prot.files)){
#   temp.l <- as.data.frame(assays(get(paste0("prot.se.",i)))$NPX_LOD) %>% 
#     mutate(feature=rownames(.)) %>% 
#     gather(key =  "ID", value = "NPX", -feature)
#   
#   p <- ggplot(temp.l, aes(x=ID, y=NPX))
#   p + geom_boxplot(aes(fill=ID))+ 
#     coord_flip()+
#     ggtitle("NPX (LOD)")+
#     theme(legend.position="none")
#   print(p)
# }
```

## PCA
```{r message=TRUE, warning=TRUE}
for(i in 1:length(prot.files)){
  
  tmp <- assays(get(paste0("prot.se.",i)))$NPX_LOD
  if(any(apply(tmp,1, var)==0)){
    tmp <- tmp[apply(tmp,1, var)!=0,]
    message("Some feature have zero variance, these have been removed")
  }
  #colnames(tmp) <- paste0(get(paste0("prot.se.",i))$ID, "_", get(paste0("prot.se.",i))$VISITnumber)
  #prcomp(t(tmp), center = T, scale.=F) -> tmp
  
  PCA(t(tmp), scale.unit = F, graph = F, ncp = nrow(tmp)) -> tmp
  
  p1 <- fviz_pca_ind(tmp,
                     col.ind = get(paste0("prot.se.",i))$Chip.name,
                     geom.ind = "point", mean.point=F, pointshape=19)+
    ggtitle(metadata(get(paste0("prot.se.",i)))$panel)
  
  p2 <- fviz_pca_ind(tmp,
                     col.ind =get(paste0("prot.se.",i))$Chip.name,
                     geom.ind = c("point", "text"), mean.point=F, pointshape=19)+
    ggtitle(metadata(get(paste0("prot.se.",i)))$panel)
  
  # p3 <- fviz_pca_ind(tmp,
  #                    col.ind = get(paste0("prot.se.",i))$Chip.name,
  #                    geom.ind = "point", mean.point=F, pointshape=19)+
  #   geom_text(aes(x=tmp$ind$coord[,1],
  #                 y=tmp$ind$coord[,2],
  #                 col = get(paste0("prot.se.",i))$Chip.name),
  #             label=get(paste0("prot.se.",i))$IID)+
  #   ggtitle(metadata(get(paste0("prot.se.",i)))$panel)
  
  p4 <- fviz_pca_ind(tmp,
                     col.ind = get(paste0("prot.se.",i))$Flag,
                     geom.ind = "point", mean.point=F, pointshape=19)+
    ggtitle(metadata(get(paste0("prot.se.",i)))$panel)
  
  p5 <- fviz_pca_ind(tmp,
                     col.ind = get(paste0("prot.se.",i))$Sex,
                     geom.ind = "point", mean.point=F, pointshape=19)+
    ggtitle(metadata(get(paste0("prot.se.",i)))$panel)
  
  print(p1)
  print(p2)
  #print(p3)
  print(p4)
  
  message(paste0("Done with ", metadata(get(paste0("prot.se.",i)))$panel))
}

```

For the CVDII Inflammation and OncologyII panels, some problems are detected. I will investigate what are the protein contribution.

```{r results="asis"}

# CVDII
tmp <- assays(prot.se.2)$NPX_LOD
  if(any(apply(tmp,1, var)==0)){
    tmp <- tmp[apply(tmp,1, var)!=0,]
    message("Some feature have zero variance, these have been removed")
  }
PCA(t(tmp), scale.unit = F, graph = F, ncp = nrow(tmp)) -> tmp

DT::datatable(as.data.frame(tmp$eig),
              options = list(autoWidth = TRUE))

DT::datatable(as.data.frame(tmp$var$contrib),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(tmp$var$cos2),
              options = list(autoWidth = TRUE, scrollX=TRUE))

fviz_contrib(tmp, choice = "var", axes=1)
fviz_contrib(tmp, choice = "var", axes=2)

```

```{r results="asis"}

# Inflammation
tmp <- assays(prot.se.4)$NPX_LOD
  if(any(apply(tmp,1, var)==0)){
    tmp <- tmp[apply(tmp,1, var)!=0,]
    message("Some feature have zero variance, these have been removed")
  }
PCA(t(tmp), scale.unit = F, graph = F, ncp = nrow(tmp)) -> tmp

DT::datatable(as.data.frame(tmp$eig),
              options = list(autoWidth = TRUE))

DT::datatable(as.data.frame(tmp$var$contrib),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(tmp$var$cos2),
              options = list(autoWidth = TRUE, scrollX=TRUE))

fviz_contrib(tmp, choice = "var", axes=1)
fviz_contrib(tmp, choice = "var", axes=2)

```
```{r results="asis"}

# OncologyII
tmp <- assays(prot.se.6)$NPX_LOD
  if(any(apply(tmp,1, var)==0)){
    tmp <- tmp[apply(tmp,1, var)!=0,]
    message("Some feature have zero variance, these have been removed")
  }
PCA(t(tmp), scale.unit = F, graph = F, ncp = nrow(tmp)) -> tmp

DT::datatable(as.data.frame(tmp$eig),
              options = list(autoWidth = TRUE))

DT::datatable(as.data.frame(tmp$var$contrib),
              options = list(autoWidth = TRUE, scrollX=TRUE))

DT::datatable(as.data.frame(tmp$var$cos2),
              options = list(autoWidth = TRUE, scrollX=TRUE))

fviz_contrib(tmp, choice = "var", axes=1)
fviz_contrib(tmp, choice = "var", axes=2)

```
The PCA score plots seems to be originating from a few outliers proteins:

1. CVDII panel: 142_FS (Uniprot:P19883, GeneSymbol:FST, GeneName: Follistatin)
2. Inflammation panel: 103_BDNF (Uniprot:P23560, GeneSymbol:BDNF, GeneName: Brain-derived neurotrophic factor)
3. OncologyII panel: 188_FR-gamma  (Uniprot:P41439, GeneSymbol:FOLR3, GeneName: Folate receptor gamma)

1 seems to be a technical outlier, as the samples belonging to the very same chip have similar negative score in PC1. In fact all samples in plate 3 of the CVDII panel have missing data for the 142_FS protein. However, 2 and 3 do not seem to be attributable to a chip effect. It remain to be investigated if they reflect true biological signal. A plot is available below

CVDII panel: 142_FS

```{r}
data.frame(NPX_LOD=assays(prot.se.2)$NPX_LOD["X142_FS",], ID=colnames(prot.se.2), Chip=prot.se.2$Chip.name) %>%
  arrange(Chip) %>% 
  ggplot(aes(x=1:ncol(prot.se.2), y=NPX_LOD))+
  geom_segment(aes(col=Chip, xend=1:ncol(prot.se.2), yend=0), stat = "identity", size=1)+
  ggtitle("142_FS")
```
Inflammation panel: 103_BDNF

```{r}
data.frame(NPX_LOD=assays(prot.se.4)$NPX_LOD["X103_BDNF",], ID=colnames(prot.se.2), Chip=prot.se.2$Chip.name) %>%
  arrange(Chip) %>% 
  ggplot(aes(x=1:ncol(prot.se.4), y=NPX_LOD))+
  geom_segment(aes(col=Chip, xend=1:ncol(prot.se.2), yend=0), stat = "identity", size=1)+
  ggtitle("103_BDNF")
```

OncologyII panel: 188_FR-gamma

```{r}
data.frame(NPX_LOD=assays(prot.se.6)$NPX_LOD["X188_FR.gamma",], ID=colnames(prot.se.2), Chip=prot.se.2$Chip.name) %>%
  arrange(Chip) %>% 
  ggplot(aes(x=1:ncol(prot.se.6), y=NPX_LOD))+
  geom_segment(aes(col=Chip, xend=1:ncol(prot.se.2), yend=0), stat = "identity", size=1)+
  ggtitle("188_FR-gamma")
```
# Feature and sample CV

```{r}
for(i in 1:length(prot.files)){
  
  tmp <- assays(get(paste0("prot.se.",i)))$NPX_LOD
  
  cv.f <- data.frame(CV=apply(tmp,1, function(x) sd(2^x, na.rm = T)/mean(2^x,na.rm=T)),
                     feature=rownames(tmp))
  cv.s <- data.frame(CV=apply(tmp,2, function(x) sd(2^x, na.rm = T)/mean(2^x,na.rm=T)),
                     ID=colnames(tmp),
                     Chip=get(paste0("prot.se.",i))$Chip.name,
                     Flag=get(paste0("prot.se.",i))$Flag)
  
  p1 <- cv.f %>%
    arrange(desc(CV)) %>%
    ggplot(aes(x=1:nrow(tmp), y=CV))+
    geom_segment(aes(xend=1:nrow(tmp), yend=0), stat = "identity", size=1)+
    scale_x_continuous("feature index", labels = cv.f$feature[order(cv.f$CV, decreasing = T)], breaks = 1:nrow(tmp))+
    ggtitle(paste0("Feature CV - ",metadata(get(paste0("prot.se.",i)))$panel))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  
  print(p1)
  
  p2 <- cv.s %>% 
    arrange(Chip) %>% 
    ggplot(aes(x=1:ncol(tmp), y=CV))+
    geom_segment(aes(col=Chip, xend=1:ncol(tmp), yend=0), stat = "identity", size=1)+
    geom_point(aes(shape=Flag))+
    scale_shape_manual(values=c("Passed"=NA,"Flagged"=19))+
    xlab("sample index")+
    ggtitle(paste0("Sample CV - ",metadata(get(paste0("prot.se.",i)))$panel))+
     theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  
  print(p2)
  
}
```

# Fraction of features and samples with NA

```{r}
for(i in 1:length(prot.files)){
  
  tmp <- assays(get(paste0("prot.se.",i)))$NPX_NaN
  
  na.f <- data.frame(n.miss=apply(tmp,1, function(x) sum(is.na(x))/length(x)),
                     feature=rownames(tmp))
  na.s <- data.frame(n.miss=apply(tmp,2, function(x) sum(is.na(x))/length(x)),
                     ID=colnames(tmp),
                     Chip=get(paste0("prot.se.",i))$Chip.name,
                     Flag=get(paste0("prot.se.",i))$Flag)
  
  p1 <- na.f %>%
    arrange(desc(n.miss)) %>%
    ggplot(aes(x=1:nrow(tmp), y=n.miss))+
    geom_segment(aes(xend=1:nrow(tmp), yend=0), stat = "identity", size=1)+
    scale_x_continuous("feature index", labels = na.f$feature[order(na.f$n.miss, decreasing = T)], breaks = 1:nrow(tmp))+
    scale_y_continuous(limits = c(0,1))+
    ggtitle(paste0("Numer of missing data - ",metadata(get(paste0("prot.se.",i)))$panel))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  
  print(p1)
  
  p2 <- na.s %>% 
    arrange(Chip) %>% 
    ggplot(aes(x=1:ncol(tmp), y=n.miss))+
    geom_segment(aes(col=Chip, xend=1:ncol(tmp), yend=0), stat = "identity", size=1)+
    geom_point(aes(shape=Flag))+
    scale_shape_manual(values=c("Passed"=NA,"Flagged"=19))+
    xlab("sample index")+
    scale_y_continuous(limits = c(0,1))+
    ggtitle(paste0("Numer of missing data - ",metadata(get(paste0("prot.se.",i)))$panel))+
     theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  
  print(p2)
  
}
```


# Intraclass Correlation Coefficient
Calculate Intraclass Correlation Coefficient (ICC) to score the normalization methods. In principle, samples belonging to the same subject should be more similar to each other.

```{r, message=F, warning=F}
ICC.l <- list()
for(i in 1:length(prot.files)){
  
  ICC.l[[i]] <- matrix(nrow=nrow(get(paste0("prot.se.",i))), ncol = 1,
                       dimnames = list(rownames(get(paste0("prot.se.",i))), metadata(get(paste0("prot.se.",i)))$panel))
  for(j in 1:nrow(ICC.l[[i]])){
    trash <- capture.output(ICC.l[[i]][j,] <- ICCbare(IID,NPX_LOD,
                                                      data=data.frame(IID=get(paste0("prot.se.",i))$IID,
                                                                      NPX_LOD=assays(get(paste0("prot.se.",i)))$NPX_LOD[j,])))
  }
  names(ICC.l)[i] <- metadata(get(paste0("prot.se.",i)))$panel
}

boxplot(ICC.l, las=2)

as.data.frame(sapply(ICC.l, median))

as.data.frame(sapply(ICC.l, quantile))

```