---
title: "DHR LCMS v03"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 7
    fig_height: 7
    theme: cerulean
    toc: yes
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

Require all the libraries
```{r}
require(SummarizedExperiment)
require(tidyverse)
require(openxlsx)
require(plotly)
require(limma)
require(geneplotter)
require(factoextra)
require(impute)
require(sva)
require(ruv)
require(corrplot)
require(RColorBrewer)
require(ICC)
```

# Read the data

Read the data and store as SummarizedExperiment. Create two separate objects for negative and positive mode data.

```{r}
outDir <- "/home/francesco.marabita/Projects/DHR/data/Metabolomics/out/LCMS"
#dir.create(outDir)

# read data
assayData.neg <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "assayData-neg", colNames =T, rowNames = T, skipEmptyRows=T)
assayData.pos <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "assayData-pos", colNames =T, rowNames = T, skipEmptyRows=T)

fData.neg <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "fData-neg", colNames =T, rowNames = F, skipEmptyRows=T)
rownames(fData.neg) <- fData.neg$feature.ID
fData.pos <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "fData-pos", colNames =T, rowNames = F, skipEmptyRows=T)
rownames(fData.pos) <- fData.pos$feature.ID

pData.neg_ <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "pData-neg", colNames =T, rowNames = F, skipEmptyRows=T)
pData.pos_ <- read.xlsx("/home/francesco.marabita/Projects/DHR/data/Metabolomics/All_LCMS_2018-11-09.xlsx",
                       sheet = "pData-pos", colNames =T, rowNames = F, skipEmptyRows=T)

# add 
load("/home/francesco.marabita/Projects/DHR/data/Clinical/Clin_data_2018-02-13.RData")

tmp <- as.data.frame(clin.data[, c("ID", "visit_id", "VISITnumber", "dateofbirth", "Age", "Visit", "Visittime", "Sex")])
colnames(tmp) <- c("IID", "ID", "Visit.number", "Birth.date", "Age", "Visit.date", "Visit.time", "Sex")

pData.neg <- merge(pData.neg_, tmp, by=c("ID", "IID", "Visit.number"), all.x=T, sort=F)
pData.neg <- pData.neg[match(pData.neg_$ID, pData.neg$ID),]
rownames(pData.neg) <- pData.neg$ID

pData.pos <- merge(pData.pos_, tmp, by=c("ID", "IID", "Visit.number"), all.x=T, sort=F)
pData.pos <- pData.pos[match(pData.pos_$ID, pData.pos$ID),]
rownames(pData.pos) <- pData.pos$ID

if(all.equal(pData.neg, pData.pos)) {
  pData <- pData.neg
}  else { stop("pData data.frames are different") }

# create SummarizedExperiment
LCMS.neg.se <- SummarizedExperiment(assays = list(RC=as.matrix(assayData.neg)))
LCMS.pos.se <- SummarizedExperiment(assays = list(RC=as.matrix(assayData.pos)))

if(all(rownames(LCMS.neg.se)==rownames(fData.neg)) & all(colnames(LCMS.neg.se)==rownames(pData))) {
  rowData(LCMS.neg.se) <- DataFrame(fData.neg)
  colData(LCMS.neg.se) <- DataFrame(pData)
} else { stop("Order is different") }

if(all(rownames(LCMS.pos.se)==rownames(fData.pos)) & all(colnames(LCMS.pos.se)==rownames(pData))) {
  rowData(LCMS.pos.se) <- DataFrame(fData.pos)
  colData(LCMS.pos.se) <- DataFrame(pData)
} else { stop("Order is different") }


# include only subjects present in clin.data
LCMS.neg.se <- LCMS.neg.se[,LCMS.neg.se$IID %in% unique(clin.data$ID)]
LCMS.pos.se <- LCMS.pos.se[,LCMS.pos.se$IID %in% unique(clin.data$ID)]

# remove samples in LCMS-pos that failed completely (100% NAs)
LCMS.pos.se <- LCMS.pos.se[,!colSums(is.na(assays(LCMS.pos.se)$RC))==nrow(LCMS.pos.se)]
```

 Contrary to GCMS data, there are a lot of missing values.

```{r}
plot(LCMS.neg.se$Runorder, colSums(is.na(assays(LCMS.neg.se)$RC))/nrow(LCMS.neg.se),
     main="Fraction of NAs per sample - LCMS-neg", xlab="Run order", ylab="Fraction of NAs")
plot(LCMS.pos.se$Runorder, colSums(is.na(assays(LCMS.pos.se)$RC))/nrow(LCMS.pos.se),
     main="Fraction of NAs per sample - LCMS-pos", xlab="Run order", ylab="Fraction of NAs")

barplot(rowSums(is.na(assays(LCMS.neg.se)$RC))/ncol(LCMS.neg.se),
         main="Fraction of NAs per feature - LCMS-neg")
barplot(rowSums(is.na(assays(LCMS.pos.se)$RC))/ncol(LCMS.pos.se),
         main="Fraction of NAs per feature - LCMS-pos")
```
I will consider only identified metabilited detected in at least 75% of the subjects.

```{r}
# Remove unidentified LCMS features
feat.out <- rownames(LCMS.neg.se)[grepl("^Peak_", rowData(LCMS.neg.se)$Metabolite.name)]
LCMS.neg.se <- LCMS.neg.se[!rownames(LCMS.neg.se) %in% feat.out,]
feat.out <- rownames(LCMS.pos.se)[grepl("^Peak_", rowData(LCMS.pos.se)$Metabolite.name)]
LCMS.pos.se <- LCMS.pos.se[!rownames(LCMS.pos.se) %in% feat.out,]

# Remove features with > 75% NAs
LCMS.neg.se <- LCMS.neg.se[rowSums(!is.na(assays(LCMS.neg.se)$RC)) > (0.75*ncol(LCMS.neg.se)),]
LCMS.pos.se <- LCMS.pos.se[rowSums(!is.na(assays(LCMS.pos.se)$RC)) > (0.75*ncol(LCMS.pos.se)),]

plot(LCMS.neg.se$Runorder, colSums(is.na(assays(LCMS.neg.se)$RC))/nrow(LCMS.neg.se),
     main="Fraction of NAs per sample - LCMS-neg", xlab="Run order", ylab="Fraction of NAs")
plot(LCMS.pos.se$Runorder, colSums(is.na(assays(LCMS.pos.se)$RC))/nrow(LCMS.pos.se),
     main="Fraction of NAs per sample - LCMS-pos", xlab="Run order", ylab="Fraction of NAs")

barplot(rowSums(is.na(assays(LCMS.neg.se)$RC))/ncol(LCMS.neg.se),
         main="Fraction of NAs per feature - LCMS-neg")
barplot(rowSums(is.na(assays(LCMS.pos.se)$RC))/ncol(LCMS.pos.se),
         main="Fraction of NAs per feature - LCMS-pos")
```

Convert the Relative Concentration (RC) values top log2(RC) and impute missing values with KNN.

```{r}
assays(LCMS.neg.se)$log2RC <- log2(assays(LCMS.neg.se)$RC)
assays(LCMS.neg.se)$log2RC <- impute.knn(assays(LCMS.neg.se)$log2RC, maxp = nrow(LCMS.neg.se)+1)$data
assays(LCMS.pos.se)$log2RC <- log2(assays(LCMS.pos.se)$RC)
assays(LCMS.pos.se)$log2RC <- impute.knn(assays(LCMS.pos.se)$log2RC, maxp = nrow(LCMS.pos.se)+1)$data
```

Check the time drift for IS and metabolites.

```{r}
as.data.frame(assays(LCMS.neg.se)$log2RC) %>% 
  mutate(feature=rownames(.), type=ifelse(grepl("IS", rownames(.),fixed = T), "IS", "metabolite")) %>% 
  gather(key =  "ID", value = "log2RC", -(feature:type)) %>% 
  group_by(feature) %>% 
  mutate(z.score.feature = (log2RC - mean(log2RC)) / sd(log2RC))-> temp.l
temp.l$Runorder <- LCMS.neg.se$Runorder[match(temp.l$ID, LCMS.neg.se$ID)]

subset(temp.l, type=="IS") %>% 
ggplot(aes(x=Runorder, y=log2RC)) +
  geom_point(aes(col=feature), alpha=0.5) +
  geom_smooth(aes(col=feature), method="loess")

subset(temp.l, type=="IS") %>% 
ggplot(aes(x=Runorder, y=z.score.feature)) +
  geom_point(aes(col=feature), alpha=0.5) +
  geom_smooth(aes(col=feature), method="loess")

ggplot(temp.l, aes(x=Runorder, y=z.score.feature)) +
  geom_smooth()


as.data.frame(assays(LCMS.pos.se)$log2RC) %>% 
  mutate(feature=rownames(.), type=ifelse(grepl("IS", rownames(.),fixed = T), "IS", "metabolite")) %>% 
  gather(key =  "ID", value = "log2RC", -(feature:type)) %>%
  group_by(feature) %>% 
  mutate(z.score.feature = (log2RC - mean(log2RC)) / sd(log2RC))-> temp.l
temp.l$Runorder <- LCMS.pos.se$Runorder[match(temp.l$ID, LCMS.pos.se$ID)]

subset(temp.l, type=="IS") %>% 
ggplot(aes(x=Runorder, y=log2RC)) +
  geom_point(aes(col=feature), alpha=0.5) +
  geom_smooth(aes(col=feature), method="loess")

subset(temp.l, type=="IS") %>% 
ggplot(aes(x=Runorder, y=z.score.feature)) +
  geom_point(aes(col=feature), alpha=0.5) +
  geom_smooth(aes(col=feature), method="loess")

ggplot(temp.l, aes(x=Runorder, y=z.score.feature)) +
  geom_smooth()


```


# Quality control
In order to check for sample-to-sample relationship, outliers and batch effects, I consider the following:

* Density plot
* Boxpots
* PCA
* Pairwise scatterplots

```{r}
multidensity(assays(LCMS.neg.se)$log2RC, legend = F, main="LCMS-neg",
             xlab="log2 Relative Concentration", xlim=c(0,30))
multidensity(assays(LCMS.pos.se)$log2RC, legend = F, main="LCMS-pos",
             xlab="log2 Relative Concentration", xlim=c(0,30))
```


```{r  fig.height=20}
temp.l <- as.data.frame(assays(LCMS.neg.se)$log2RC) %>% 
  mutate(feature=rownames(.)) %>% 
  gather(key =  "ID", value = "log2RC", -feature)

p <- ggplot(temp.l, aes(x=ID, y=log2RC))
p + geom_boxplot(aes(fill=ID))+ 
  coord_flip()+
  ggtitle("log2RC - LCMS-neg")+
  theme(legend.position="none")

temp.l <- as.data.frame(assays(LCMS.pos.se)$log2RC) %>% 
  mutate(feature=rownames(.)) %>% 
  gather(key =  "ID", value = "log2RC", -feature)

p <- ggplot(temp.l, aes(x=ID, y=log2RC))
p + geom_boxplot(aes(fill=ID))+ 
  coord_flip()+
  ggtitle("log2RC - LCMS-pos")+
  theme(legend.position="none")
```

Finally, I make a PCA plot, coloring by the runorder  (`Runorder`), as I do not have or the experimental batch (`Analytical.Batch`) or the runorder time in minutes (`Exp.Time.Difference.minutes`), similarly to the GCMS data.
```{r}
temp <- t(assays(LCMS.neg.se)$log2RC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.neg.se$Runorder),
            label=LCMS.neg.se$IID)


temp <- t(assays(LCMS.pos.se)$log2RC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.pos.se$Runorder),
            label=LCMS.pos.se$IID)

rm(temp)
```
# Normalization
I will compare Quantile Normalization, Median Scaling, RUV4. I do not have a categorical batch variable (not annotated), therefore I will not evaluate ComBat

## Quantile Normalization
```{r}
assays(LCMS.neg.se)$log2QNRC <- normalizeQuantiles(assays(LCMS.neg.se)$log2RC)
assays(LCMS.pos.se)$log2QNRC <- normalizeQuantiles(assays(LCMS.pos.se)$log2RC)
```


```{r}
multidensity(assays(LCMS.neg.se)$log2QNRC, legend = F, main="Quantile Norm",
             xlab="Relative concentration", xlim=c(0,40))
multidensity(assays(LCMS.pos.se)$log2QNRC, legend = F, main="Quantile Norm",
             xlab="Relative concentration", xlim=c(0,40))

```

```{r}
temp <- t(assays(LCMS.neg.se)$log2QNRC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.neg.se$Runorder),
            label=LCMS.neg.se$IID)


temp <- t(assays(LCMS.pos.se)$log2QNRC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.pos.se$Runorder),
            label=LCMS.pos.se$IID)

rm(temp)
```


## Median Scaling
```{r}
assays(LCMS.neg.se)$log2NRC <- normalizeBetweenArrays(assays(LCMS.neg.se)$log2RC, method = "scale")
assays(LCMS.pos.se)$log2NRC <- normalizeBetweenArrays(assays(LCMS.pos.se)$log2RC, method = "scale")
```


```{r}
multidensity(assays(LCMS.neg.se)$log2NRC, legend = F, main="Median Scaling",
             xlab="Relative concentration", xlim=c(0,40))
multidensity(assays(LCMS.pos.se)$log2NRC, legend = F, main="Median Scaling",
             xlab="Relative concentration", xlim=c(0,40))

```

```{r}
temp <- t(assays(LCMS.neg.se)$log2NRC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.neg.se$Runorder),
            label=LCMS.neg.se$IID)


temp <- t(assays(LCMS.pos.se)$log2NRC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.pos.se$Runorder),
            label=LCMS.pos.se$IID)

rm(temp)
```

## RUV4

With RUV4, I will use the Internal Standards (ISs) as negative control metabolites, with the assumption that they are not influenced by biological variation, but any observed variability in the IS is enitirelly introduced byt the experimental steps. The method requires the coiche of an optimal number of K, which is not trivial. RUV4 however is less sensitive to the number of Ks as compared to RUV2. I will evaluate the correlation of the unobserved covariates W with the control genes, as well as the distribution of the relative concentrations. The method requires to specify a design matrix, and for this purpose I use the individual (`IID`) and the visit numer (`Visit.number`) as covariate. I first get a rough estimation of the Ks with the "leave1out" method.

```{r}
numk <- getK(Y = t(assays(LCMS.neg.se)$log2RC) ,
             X = cbind(LCMS.neg.se$IID,as.factor(LCMS.neg.se$Visit.number)),
             ctl = grepl("_IS_", rownames(LCMS.neg.se), fixed=T),
             method = "leave1out")

numk$k

numk <- getK(Y = t(assays(LCMS.pos.se)$log2RC) ,
             X = cbind(LCMS.pos.se$IID,as.factor(LCMS.pos.se$Visit.number)),
             ctl = grepl("_IS_", rownames(LCMS.pos.se), fixed=T),
             method = "leave1out")
numk$k
```

```{r}
fit.neg <- RUV4(Y = t(assays(LCMS.neg.se)$log2RC),
                X = cbind(LCMS.neg.se$IID, as.factor(LCMS.neg.se$Visit.number)),
                ctl = grepl("_IS_", rownames(LCMS.neg.se), fixed=T),
                k=2)
fit.pos <- RUV4(Y = t(assays(LCMS.pos.se)$log2RC),
                X = cbind(LCMS.pos.se$IID, as.factor(LCMS.pos.se$Visit.number)),
                ctl = grepl("_IS_", rownames(LCMS.pos.se), fixed=T),
                k=1)

```

The normalized conncentrations are then obtained by removing the effect of the W components. From the RUV4 documentation, `W`	represent the estimated unwanted factors and `alpha`is the estimated coefficients of `W`.

```{r}
assays(LCMS.neg.se)$ruv4RC <- t(t(assays(LCMS.neg.se)$log2RC) - fit.neg$W %*% fit.neg$alpha)
assays(LCMS.pos.se)$ruv4RC <- t(t(assays(LCMS.pos.se)$log2RC) - fit.pos$W %*% fit.pos$alpha)
```


```{r}
temp <- t(assays(LCMS.neg.se)$ruv4RC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.neg.se$Runorder),
            label=LCMS.neg.se$IID)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.neg.se$Sex, axes = c(1,2),
             geom.ind = "point", mean.point=F, pointshape=19)

temp <- t(assays(LCMS.pos.se)$ruv4RC)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col = LCMS.pos.se$Runorder),
            label=LCMS.pos.se$IID)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = LCMS.pos.se$Sex, axes = c(1,2),
             geom.ind = "point", mean.point=F, pointshape=19)

rm(temp)
```
```{r}
data.frame(Runorder=LCMS.neg.se$Runorder, as.data.frame(fit.neg$W)) %>%
  gather(key = "W", value="value", -Runorder) %>% 
  ggplot(aes(x=Runorder, y=value))+
  geom_point(aes(col=Runorder))+
  facet_wrap(~ W, scales="free")

data.frame(Runorder=LCMS.pos.se$Runorder, as.data.frame(fit.pos$W)) %>%
  gather(key = "W", value="value", -Runorder) %>% 
  ggplot(aes(x=Runorder, y=value))+
  geom_point(aes(col=Runorder))+
  facet_wrap(~ W, scales="free")
```

```{r}
data.frame(as.data.frame(t(assays(LCMS.neg.se[grepl("_IS_", rownames(LCMS.neg.se), fixed=T),])$log2RC)),
           Runorder=LCMS.neg.se$Runorder) %>%
  mutate(type="log2RC") %>% 
  gather(key="IS", value="value", -c(Runorder:type)) -> temp1

data.frame(as.data.frame(t(assays(LCMS.neg.se[grepl("_IS_", rownames(LCMS.neg.se), fixed=T),])$ruv4RC)),
           Runorder=LCMS.neg.se$Runorder) %>%
  mutate(type="ruv4RC") %>% 
  gather(key="IS", value="value", -c(Runorder:type)) -> temp2

temp <- rbind(temp1, temp2)

ggplot(temp, aes(x=Runorder, y=value))+
  geom_point(aes(col=type), alpha=0.3)+
  geom_smooth(aes(group=type, lty=type), se = T, method = "lm")+
  facet_wrap(~ IS, scale="fixed")


data.frame(as.data.frame(t(assays(LCMS.pos.se[grepl("_IS_", rownames(LCMS.pos.se), fixed=T),])$log2RC)),
           Runorder=LCMS.pos.se$Runorder) %>%
  mutate(type="log2RC") %>% 
  gather(key="IS", value="value", -c(Runorder:type)) -> temp1

data.frame(as.data.frame(t(assays(LCMS.pos.se[grepl("_IS_", rownames(LCMS.pos.se), fixed=T),])$ruv4RC)),
           Runorder=LCMS.pos.se$Runorder) %>%
  mutate(type="ruv4RC") %>% 
  gather(key="IS", value="value", -c(Runorder:type)) -> temp2

temp <- rbind(temp1, temp2)

ggplot(temp, aes(x=Runorder, y=value))+
  geom_point(aes(col=type), alpha=0.3)+
  geom_smooth(aes(group=type, lty=type), se = T, method = "lm")+
  facet_wrap(~ IS, scale="fixed")

```

# Comparison

## Reduction of concentration drift over time

A boxplot of the slopes will reveal any reduction in the linear trend, for the ISs and the metabolites.

```{r}
time.trend.neg <- list()
methods <- names(assays(LCMS.neg.se))
methods <- methods[-1]
names(methods) <- c("log2RC", "log2QNRC","log2NRC","RUV4")

x <- LCMS.neg.se$Runorder
is <- rownames(LCMS.neg.se)[grepl("_IS_", rownames(LCMS.neg.se), fixed=T)]
for(i in 1:length(methods)){
  time.trend.neg[[i]] <- list()
  time.trend.neg[[i]]$slope <- apply(assays(LCMS.neg.se)[[methods[i]]], 1, function(y) coef(lm(y ~ x))["x"])
  time.trend.neg[[i]]$intercept<- apply(assays(LCMS.neg.se)[[methods[i]]], 1, function(y) coef(lm(y ~ x))["(Intercept)"])
  names(time.trend.neg)[i] <- names(methods)[i]
}

as.data.frame(lapply(time.trend.neg, "[[", "slope")) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="slope", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=slope))+
  geom_boxplot(aes(fill=type),outlier.color = NA)+
  geom_jitter(width=0.1, alpha=0.5)+
  geom_hline(yintercept = 0, lty=2, col="red")+
  scale_x_discrete(limits=names(time.trend.neg))+
  facet_wrap(~featureType,scales = "free")

as.data.frame(lapply(time.trend.neg, "[[", "intercept")) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="intercept", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=intercept))+
  geom_boxplot(aes(fill=type),outlier.color = NA)+
  geom_jitter(width=0.1, alpha=0.5)+
  scale_x_discrete(limits=names(time.trend.neg))+
  facet_wrap(~featureType,scales = "free")

time.trend.pos <- list()
x <- LCMS.pos.se$Runorder
is <- rownames(LCMS.pos.se)[grepl("_IS_", rownames(LCMS.pos.se), fixed=T)]
for(i in 1:length(methods)){
  time.trend.pos[[i]] <- list()
  time.trend.pos[[i]]$slope <- apply(assays(LCMS.pos.se)[[methods[i]]], 1, function(y) coef(lm(y ~ x))["x"])
  time.trend.pos[[i]]$intercept<- apply(assays(LCMS.pos.se)[[methods[i]]], 1, function(y) coef(lm(y ~ x))["(Intercept)"])
  names(time.trend.pos)[i] <- names(methods)[i]
}

as.data.frame(lapply(time.trend.pos, "[[", "slope")) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="slope", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=slope))+
  geom_boxplot(aes(fill=type),outlier.color = NA)+
  geom_jitter(width=0.1, alpha=0.5)+
  geom_hline(yintercept = 0, lty=2, col="red")+
  scale_x_discrete(limits=names(time.trend.pos))+
  facet_wrap(~featureType,scales = "free")

as.data.frame(lapply(time.trend.pos, "[[", "intercept")) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="intercept", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=intercept))+
  geom_boxplot(aes(fill=type),outlier.color = NA)+
  geom_jitter(width=0.1, alpha=0.5)+
  scale_x_discrete(limits=names(time.trend.pos))+
  facet_wrap(~featureType,scales = "free")


```

## Reduction of the CV

The coefficient of variation should also be decreased, assuming that part of the technical variability has been removed with the normalization methods. The CV is calculated on the linear scale.

```{r}
cv.neg <- list()
methods <- names(assays(LCMS.neg.se))
methods <- methods[-1]
names(methods) <- c("log2RC", "log2QNRC","log2NRC","RUV4")
for(i in 1:length(methods)){
  cv.neg[[i]] <- apply(assays(LCMS.neg.se)[[methods[i]]], 1, 
                       function(y) sd(2^y)/mean(2^y))
  names(cv.neg)[i] <- names(methods)[i]
}

as.data.frame(cv.neg) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="cv", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=cv))+
  geom_boxplot(aes(fill=type))+
  geom_jitter(width=0.1, alpha=0.5)+
  scale_x_discrete(limits=names(cv))+
  facet_wrap(~featureType,scales = "free")


cv.pos <- list()
methods <- names(assays(LCMS.pos.se))
methods <- methods[-1]
names(methods) <- c("log2RC", "log2QNRC","log2NRC","RUV4")
for(i in 1:length(methods)){
  cv.pos[[i]] <- apply(assays(LCMS.pos.se)[[methods[i]]], 1, 
                         function(y) sd(2^y)/mean(2^y))
  names(cv.pos)[i] <- names(methods)[i]
}

as.data.frame(cv.pos) %>% 
  mutate(feature=rownames(.)) %>% 
  mutate(featureType=(ifelse(grepl("_IS_", feature, fixed=T), "IS", "metabolite"))) %>%
  gather(key="type", value="cv", -(feature:featureType)) %>% 
  ggplot(aes(x=type, y=cv))+
  geom_boxplot(aes(fill=type))+
  geom_jitter(width=0.1, alpha=0.5)+
  scale_x_discrete(limits=names(cv))+
  facet_wrap(~featureType,scales = "free")

```

## Separation for selected metabolites

### Steroids
```{r}

stripchart(assays(LCMS.neg.se)$log2RC["LCMS_neg_feature_79",] ~ LCMS.neg.se$Sex,
           pch=19, vertical=T, method="jitter", ylab = "log2RC", main="androsterone sulfate",
           col=c("black", "red"))

stripchart(assays(LCMS.neg.se)$log2RC["LCMS_neg_feature_83",] ~ LCMS.neg.se$Sex,
           pch=19, vertical=T, method="jitter", ylab = "log2RC", main="testosterone sulfate",
           col=c("black", "red"))

plot(assays(LCMS.neg.se)$log2RC["LCMS_neg_feature_79",], col=as.numeric(LCMS.neg.se$Sex), pch=19)
plot(assays(LCMS.neg.se)$log2RC["LCMS_neg_feature_83",], col=as.numeric(LCMS.neg.se$Sex), pch=19)

stripchart(assays(LCMS.pos.se)$log2RC["LCMS_pos_feature_147",] ~ LCMS.pos.se$Sex,
           pch=19, vertical=T, method="jitter", ylab = "log2RC", main="progesterone",
           col=c("black", "red"))
stripchart(assays(LCMS.pos.se)$log2RC["LCMS_pos_feature_148",] ~ LCMS.pos.se$Sex,
           pch=19, vertical=T, method="jitter", ylab = "log2RC", main="testosterone",
           col=c("black", "red"))

plot(assays(LCMS.pos.se)$log2RC["LCMS_pos_feature_147",], col=as.numeric(LCMS.pos.se$Sex), pch=19)
plot(assays(LCMS.pos.se)$log2RC["LCMS_pos_feature_148",], col=as.numeric(LCMS.pos.se$Sex), pch=19)


methods <- names(assays(LCMS.neg.se))
methods <- methods[-1]
names(methods) <- c("log2RC", "log2QNRC","log2NRC","RUV4")

for(i in 1:length(methods)){
assays(LCMS.neg.se)[[methods[i]]][c("LCMS_neg_feature_79","LCMS_neg_feature_83"),] %>%
  as.data.frame() %>%
  mutate(feature=rownames(.)) %>% 
  gather(key="ID", value = !!methods[i], -feature) -> feat.select.neg

feat.select.neg$sex <- pData$Sex[match(feat.select.neg$ID, pData$ID)]
feat.select.neg$name <- rowData(LCMS.neg.se[feat.select.neg$feature,])$Metabolite.name

ggplot(feat.select.neg, aes_string(x="sex", y=methods[i]))+
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(aes(col=sex), alpha=0.2 ) +
  facet_wrap(~name, scales = "free_y" ) -> p

print(p)

feat.select.neg %>% 
  group_by(name) %>%
  do(broom::tidy(t.test(get(methods[i])~sex, data=.))) -> t

print(t)

}


for(i in 1:length(methods)){
assays(LCMS.pos.se)[[methods[i]]][c("LCMS_pos_feature_147","LCMS_pos_feature_148"),] %>%
  as.data.frame() %>%
  mutate(feature=rownames(.)) %>% 
  gather(key="ID", value = !!methods[i], -feature) -> feat.select.pos

feat.select.pos$sex <- pData$Sex[match(feat.select.pos$ID, pData$ID)]
feat.select.pos$name <- rowData(LCMS.pos.se[feat.select.pos$feature,])$Metabolite.name

ggplot(feat.select.pos, aes_string(x="sex", y=methods[i]))+
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(aes(col=sex), alpha=0.2 ) +
  facet_wrap(~name, scales = "free_y" ) -> p

print(p)

feat.select.pos %>% 
  group_by(name) %>%
  do(broom::tidy(t.test(get(methods[i])~sex, data=.))) -> t

print(t)

}

```

### Aminoacids associated with sex

```{r}
aa.select <- read.table("/home/francesco.marabita/Projects/DHR/data/Metabolomics/AA_sex_assoc.txt", header=T, sep="\t")

feat.select <- rownames(LCMS.pos.se)[rowData(LCMS.pos.se)$Metabolite.name %in% aa.select$Name]

methods <- names(assays(LCMS.pos.se))
methods <- methods[-1]
names(methods) <- c("log2RC", "log2QNRC","log2NRC","RUV4")

for(i in 1:length(methods)){
assays(LCMS.pos.se)[[methods[i]]][feat.select,] %>%
  as.data.frame() %>%
  mutate(feature=rownames(.)) %>% 
  gather(key="ID", value = !!methods[i], -feature) -> feat.select.l

feat.select.l$sex <- pData$Sex[match(feat.select.l$ID, pData$ID)]
feat.select.l$name <- rowData(LCMS.pos.se[feat.select.l$feature,])$Metabolite.name

ggplot(feat.select.l, aes_string(x="sex", y=methods[i]))+
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(aes(col=sex), alpha=0.2 ) +
  facet_wrap(~name, scales = "free_y" ) -> p

print(p)

feat.select.l %>% 
  group_by(name) %>%
  do(broom::tidy(t.test(get(methods[i])~sex, data=.))) -> t

print(t)

}

```

# Clustering for combined dataset

```{r}
m1 <- as.data.frame(assays(LCMS.neg.se)$ruv4RC)
m1$feature.ID <- rownames(LCMS.neg.se)
m2 <- as.data.frame(assays(LCMS.pos.se)$ruv4RC)
m2$feature.ID <- rownames(LCMS.pos.se)

LCMS.neg.pos <- merge(m1, m2, all=T, sort=F)
rownames(LCMS.neg.pos) <- LCMS.neg.pos$feature.ID
LCMS.neg.pos <- LCMS.neg.pos[, !colnames(LCMS.neg.pos) %in% "feature.ID"]


pData.neg.pos <- pData[match(colnames(LCMS.neg.pos), pData$ID),]
fData.neg.pos_ <- rbind(fData.neg, fData.pos)
fData.neg.pos <- fData.neg.pos_[match(rownames(LCMS.neg.pos), fData.neg.pos_$feature.ID),]
rm(fData.neg.pos_)

LCMS.neg.pos.se <- SummarizedExperiment(assays = list(ruv4RC=as.matrix(LCMS.neg.pos)))
if(all(rownames(LCMS.neg.pos.se)==rownames(fData.neg.pos)) & all(colnames(LCMS.neg.pos.se)==rownames(pData.neg.pos))) {
  rowData(LCMS.neg.pos.se) <- DataFrame(fData.neg.pos)
  colData(LCMS.neg.pos.se) <- DataFrame(pData.neg.pos)
} else { stop("Order is different") }

```

```{r}
temp.se <- LCMS.neg.pos.se[,colSums(is.na(assays(LCMS.neg.pos.se)$ruv4RC))==0]
temp <- t(assays(temp.se)$ruv4RC) 
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = temp.se$Runorder,
             geom.ind = "point", mean.point=F)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = temp.se$Runorder,
             geom.ind = c("point", "text"))
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             cocol.ind = temp.se$Runorder,
             geom.ind = "point")+
  geom_text(aes(x=prcomp(temp, center=T, scale.=F)$x[,1],
                y=prcomp(temp, center=T, scale.=F)$x[,2],
                col= temp.se$Runorder),
            label=temp.se$IID)
fviz_pca_ind(prcomp(temp, center=T, scale.=F),
             col.ind = temp.se$Sex, axes = c(1,2),
             geom.ind = "point", mean.point=F, pointshape=19)

```

```{r}
# save(LCMS.neg.se, LCMS.pos.se, file="/home/francesco.marabita/Projects/DHR/data/Metabolomics/out/LCMS/DHR_LCMS_se_v03.RData")
# save.image(file="/home/francesco.marabita/Projects/DHR/data/Metabolomics/out/LCMS/DHR_LCMS_v03.RData")
```

