---
title: "DHR Variance Partition v03"
author: "Francesco Marabita"
output:
  html_document:
    fig_width: 7
    fig_height: 7
    theme: cerulean
    toc: yes
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# Variance partition

I will fit a LME model to all the data to partition the variance into the *Fixed*, *Random* and *Residual* components. The *Fixed* component represents the effect of Age, Sex, and time, while the *Random* component is the between-individual effect, specified as a random intercept. 

## Load the library and the data

```{r}
require(MultiAssayExperiment)
require(SummarizedExperiment)
require(lme4)
require(insight)
require(tidyverse)

load("/home/francesco.marabita/Projects/DHR/data/MultiAssayExp/out/DHR_MAexp_v05.RData")
```

## Prepare data for linear model

* Make sure there are no feature with zero variance
* Make longFormat with covariates

```{r}
# Get rid of rows that have zero variance
for (i in 1:length(experiments(MAexp.clean))){
  experiments(MAexp.clean)[[i]] <- experiments(MAexp.clean)[[i]][apply(assay(MAexp.clean[[i]]), 1, function(x) var(x,na.rm = T) != 0), ]
}

# inlcude only omics data
data.lmm <-  longFormat(MAexp.clean, colDataCols="Sex")
#data.lmm <-  longFormat(MAexp.clean, colDataCols=c("Sex", "clust.mf.comb"))
data.lmm$Sex <- fct_relevel(data.lmm$Sex, "F", "M")
#data.lmm$group <- fct_recode(data.lmm$clust.mf.comb, Low = "F1", High = "F2", High="F3", High="F4", Low="M1", High="M2")
data.lmm$IID <- as.factor(data.lmm$primary)
data.lmm$Age <- pData$Age[match(as.character(data.lmm$colname), pData$ID)]
data.lmm$Visit.number <- as.factor(pData$Visit.number[match(as.character(data.lmm$colname), pData$ID)])
data.lmm$feature.ID <- paste0(data.lmm$assay,"_",data.lmm$rowname)
```

Check if time points can be taken approximately as equally spaced observations and create the new variable `Time.point`.

```{r}
# time points are approximately equally distributed
ggplot(as.data.frame(pData), aes(y=IID, x=Visit.date))+
  geom_point()

# consider Time.point as numeric (Visit.number - 1)
length(unique(data.lmm$IID[data.lmm$Visit.number==1])) == length(unique(data.lmm$IID))

data.lmm$Time.point <- as.numeric(as.character(data.lmm$Visit.number)) - 1
```


## Fit linear mixed  model

I will model each feature as a function of  `Age`, `Sex`, `Time.point` (0-4). A random intercept will be considered to model the subject-specific effect `(1 | IID)`.

```{r}
feature.ID <- unique(data.lmm$feature.ID)

list.lmm <- mclapply(feature.ID, function(x) {
  lmer(value ~ Age + Sex + Time.point + (1 | IID),
       data = as.data.frame(data.lmm[data.lmm$feature.ID==x,]))
}, mc.preschedule = T, mc.cores = 36)
names(list.lmm) <- feature.ID
```

## Extract variance components

```{r}
df.var <- bind_rows(lapply(list.lmm, function(x) as.data.frame(get_variance(x))), .id="feature.ID")
df.var$f.var.fixed <- df.var$var.fixed/(df.var$var.fixed + df.var$var.random + df.var$var.residual)
df.var$f.var.random <- df.var$var.random/(df.var$var.fixed + df.var$var.random + df.var$var.residual)
df.var$f.var.residual <- df.var$var.residual/(df.var$var.fixed + df.var$var.random + df.var$var.residual)

df.var %>% 
  select(feature.ID, starts_with("f.var")) %>% 
  mutate(assay=data.lmm$assay[match(feature.ID, data.lmm$feature.ID)]) %>%
  mutate(assay.group=if_else(grepl("prot\\..+$", assay),"prot",assay)) %>%
  mutate(assay.group=fct_relevel(assay.group, c("clinical","GCMS", "LCMS.neg", "LCMS.pos","autoab","prot","microb.16S"))) %>% 
  gather(key="type", value="value", -c(feature.ID, assay, assay.group)) %>% 
  ggplot(aes(type,value))+
  ggbeeswarm::geom_beeswarm(aes(col=type)) + 
  geom_boxplot(width=0.1, fill="white", outlier.colour = NA) +
  scale_x_discrete(labels=c("Fixed", "Random", "Residual")) +
  facet_wrap(~assay.group, scales = "free_x") +
  theme(legend.position = "none")
```


```{r}
df.var %>% 
  select(feature.ID, starts_with("f.var")) %>% 
  mutate(assay=data.lmm$assay[match(feature.ID, data.lmm$feature.ID)]) %>%
  mutate(assay.group=if_else(grepl("prot\\..+$", assay),"prot",assay)) %>%
  mutate(assay.group=fct_relevel(assay.group, c("clinical","GCMS", "LCMS.neg", "LCMS.pos","autoab","prot","microb.16S"))) %>% 
  gather(key="type", value="value", -c(feature.ID, assay, assay.group)) %>% 
  group_by(assay.group, type) %>% 
  summarise(average=mean(value, na.rm=T)) %>% 
  DT::datatable()
```



